---
title: "Exercise template"
author: "Wouter"
date: "7/1/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Styles to be included at the top of the document
<!-- Define programming tip style -->
<style>
.tip {
  background-color: #f5f5f5;
}
</style>

<!-- Define question style -->
<style>
.question {
  color: #5A9DDB;
}
</style>


## Code hint

More than one hint possible.

```{r exerciselabel-hint-1}

```


## Markdown (text) hint

Only one hint possible.

<div id="exerciselabel-hint">
__Hint:__ hinttext.
</div>

## Programming Tip

<div class="tip" >
__Programming Tip__

- 
</div>

## Exercise with result checker

<div class="question" >
</div>

```{r exerciselabel, exercise = TRUE}

```

```{r exerciselabel-check}
gradethis::grade_result(
  pass_if(~ identical(.result, correct_result), "message"),
  fail_if(~ identical(.result, wrong_result), "message")
)
```

## Exercise with conditions checker

<div class="question" >
</div>

```{r exerciselabel, exercise = TRUE}

```

```{r exerciselabel-check}
# every pass_if condition must be satisfied
gradethis::grade_result_strict(
  pass_if(~ identical(.result, correct_result), "message"),
  fail_if(~ identical(.result, wrong_result), "message")
)
```

## Exercise with code checker

<div class="question">
</div>

```{r exerciselabel, exercise = TRUE}

```

<!-- To hide the solution, use a textual hint. -->
<div id="exerciselabel-hint">
__Hint:__ hinttext.
</div>

```{r exerciselabel-solution}
"replace this by the correct code (pipe as one line)"
```

```{r exerciselabel-check}
gradethis::grade_code(
  correct = "message", 
  incorrect = "message"
  )
```

## Quiz

```{r quiz}
quiz(
  caption = "",
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

## Installation

- Student must install the `learnr` (`install.packages("learnr")`) and `gradethis` (`remotes::install_github("rstudio-education/gradethis"`, which will also install `learnr`) packages before a tutorial can be taken.
- Tutorials must be installed with `devtools::install_github("WdeNooy/UsingRTutorials")`. First install `devtools` package, which may require to update all other packages first.
- Updating the files on GitHub automatically updates the UsingRTutorials package: re-install/update it and you have the updated package.

##Starting a tutorial

- console: `learnr::run_tutorial("introduction", "UsingRTutorials")`; opens in browser.
- _Tutorials_ tab in RStudio: Press button _Start Tutorial_ and, when started in the tab, open in new window (cannot be opened in a browser). NOTE: once the `UsingRTutorials` package is installed, the tutorials show up in the _Tutorials_ tab (may take a few seconds), regardless of whether the tutorials package is loaded or not.

## Notes on `learnr`

- While a `learnr` tutorial is running, RStudio _Help_ and _Packages_ tabs can be used, packages can be installed in RStudio, commands can be run from the console, and it is possible to run a shiny app from RStudio. So a lot of functionality of the RStudio interface can be used. 
-	Packages cannot be installed (must be done in RStudio). 
-	Packages (libraries) loaded within a tutorial (RMarkdown file) are available in all exercises within learnr, including exercises positioned before loading the package, but not within the RStudio interface. 
-	A data set loaded in the Rmd document is available for all exercises, including exercises positioned before loading the data set.
-	A data set imported/read in one exercise (code box) cannot be used in another code box and is not present in the RStudio environment.
- It is not possible to run a shiny app from within the tutorial.
-	I can’t make an external CSS file to work, so the (programming) tip style is defined in the tutorial.

## Notes on `gradethis`

-	The provided solution is directly accessible to the student unless we add one (not more!) Markdown (text) hint. A code chunk hint is displayed as a separate answer box if there is a code chunk check.
-	Sometimes displays message “Last value being used to check answer is invisible. See `?invisible` for more information”. Tell students to ignore this message.

With grade_code(): 

-	If the correct answer does not specify argument names, argument names are considered incorrect and vice versa. Note that the “mapping =” argument does not throw an error.
- Can handle pipes and it can spot the location in a pipe where code starts to diverge as long as the correct answer is a pipe without new line characters. (Note: some gradethis builds throw an error: “Error occured while checking the submission”).
-	We cannot address the code itself in grade_result() (or any of the other gradings?).
- For a chain of setups, e.g., code-checking question 3 requires the solution to code-checking question 2, which requires the solution to code-checking question 1, both the exercise and exercise-solution chunks of code-checking question 2 must refer to the solution of code-checking question 1.
	  
With grade_result(): 

- Use a tilde if you use .result to refer to the resulting object:
    fail_if(~ ncol(.result) != 4, "Too few variables!”)
- but do not use a tilde if you use a function:
    fail_if(function(x) {ncol(x) != 4}, "Too few variables!”)
-	if you use more than one criterion, put hem between {}:
    pass_if( ~ {nrow(.result) == 13 && ncol(.result) == 4 && identical(names(.result), c("friendships", "Freq", "Perc", "CumPerc")) && round(.result[[1, 3]]) == 13  && round(.result[[2, 4]]) == 25 }, “some message”)

Comparing ggplots: 

- compare the code: any difference in the order of layers yields a fail message(?)
- compare the str (structure) output of each object:
    +	Within the –check code chunk, create the correct plot:
      a <- ggplot() + <…>
    +	Compare the user’s plot as a structure (list of strings?) to the correct plot
      pass_if( ~ diffobj::diffStr(.result, a)@diff.count.full == 0, "ok")
    +	Or: A visual solution would be to look at the final svg produced by ggplot2. vdiffr::write_svg(). If the svg values are the same, then the image is the same.  https://github.com/rstudio-education/gradethis/issues/97 
