---
title: "Session 7: Visualization 1"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Get more familiar with `ggplot()` and check out some fancy visualization stuff.
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(knitr)

gradethis::gradethis_setup()
tutorial_options(exercise.timelimit = 60)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# Ensure that library is loaded.
library(tidyverse)
```

```{r, context="data", include=FALSE}
# Ensure that the data is loaded for the remainder of this tutorial.
consumers <- UsingRTutorials::consumers #  
# Alternatives: readr::read_csv("http://82.196.4.233:3838/www/consumers.csv") or readr::read_csv("data/consumers.csv")
GlasgowFriends <- UsingRTutorials::GlasgowFriends
```

<!-- Define programming tip style -->

```{=html}
<style>
.tip {
  background-color: #f5f5f5;
}
</style>
```
<!-- Define question style -->

```{=html}
<style>
.question {
  color: #5A9DDB;
}
</style>
```
<!-- Define emphasis style -->

```{=html}
<style>
.emphasis {
  color: #e8301b;
}
</style>
```
## Overview

-   ggplot Essentials
-   Fancy graphs

## ggplot Essentials

Let's keep working on a graph we discussed in the very first session. We
used some (fake) `consumers` data about a particular brand and exposure
to an advertising campaign for the brand.

**These are the variables:**

```{r variableDesc}
tibble::tibble(
  `Variable name` = c("ad_expo", "wom", "gender", "brand_aw", "firstname"),
  `Variable Label` = c("Exposure to the campaign", "Heard about the brand through word of mouth", "Gender of the respondent", "Awareness of the brand", "Respondent's first name"),
  `Value Labels`= c("1 = No exposure; 10 = Max exposure", "yes, no", "female, male", "1 = Not aware; 10 = Max aware", "")) |>
  knitr::kable(booktabs = TRUE)
```

**Here is the plot we started to recreate in Session 1:**

```{r plot-recreate1, exercise=FALSE, message=FALSE, warning=FALSE}
# Standard ggplot plot with title and axis labels.
ggplot2::ggplot(
  data = consumers,
  mapping =
    aes(x = ad_expo,
        y = brand_aw)
  )+
  geom_point(
    mapping = aes(color = gender, shape = wom),
    size = 4
    ) +
  geom_smooth(
    method = "lm",
    formula = y ~x,
    se = FALSE,
    color = "black"
    ) +
  labs(
    title = "Does brand awareness depend on exposure, \nword-of-mouth, and gender?",
    x = "Exposure to the campaign",
    y = "Brand awareness"
    )
```

**Here is how far we got:**

```{r plot-session1Final, exercise=FALSE, message=FALSE, warning=FALSE}
ggplot(
  data = consumers,
  mapping =
    aes(x = ad_expo,
        y = brand_aw)
  )+
  geom_point(
    mapping = aes(shape = wom, color = gender),
    size = 4
    ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "black"
    )
```


### Axis Labels

::: question
Add the labels to the plot or change the labels. Mind the details.
:::

```{r ggplotLabels, exercise = TRUE, exercise.lines = 20, message=FALSE, warning=FALSE}
# The code below creates the final solution we arrived at in Session 1. 
ggplot(
  data = consumers,
  mapping =
    aes(x = ad_expo,
        y = brand_aw)
  )+
  geom_point(
    mapping = aes(shape = wom, color = gender),
    size = 4
    ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "black"
    )
```

<!-- To hide the solution, use a textual hint. -->

::: {#ggplotLabels-hint}
**Hint:** Consider the details of both axes and the title. Use '\n' to add linebreaks to a label.
:::

```{r ggplotLabels-solution}
ggplot(
  data = consumers,
  mapping =
    aes(x = ad_expo,
        y = brand_aw)
  )+
  geom_point(
    mapping = aes(shape = wom, color = gender),
    size = 4
    ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "black"
    ) +
  labs(
    title = "Does brand awareness depend on exposure, \nword-of-mouth, and gender?",
    x = "Exposure to the campaign",
    y = "Brand awareness"
    )
```

```{r ggplotLabels-check}
gradethis::grade_code()
```

### Facetting

Different graphs for different groups of observations.

-   facet_wrap(\~ x): one grouping variable
-   facet_grid(y \~ x): two grouping variables

```{r plot_facet_example, exercise=FALSE, message=FALSE, warning=FALSE}
# Adaptation: word of mouth as facets instead of shape.
ggplot(
  data = consumers,
  mapping = 
    aes(x = ad_expo,
        y =brand_aw)
  )+
  geom_point(
    mapping = aes(color = gender),
    size = 4
    ) +
  geom_smooth(
    mapping = 
    aes(color = gender),
    method = "lm",
    se = FALSE
    ) +
  labs(
    title = "Does brand awareness depend on exposure, \nword-of-mouth, and gender?",
    x = "Exposure to the campaign",
    y = "Brand awareness"
    ) +
  facet_wrap(vars(wom))
```

::: question
Adjust the visualization to match the plot above. Create two plots in
one row.
:::

```{r plot_facet, exercise=TRUE, exercise.lines = 20, warning=FALSE, message=FALSE }
# Copy the solution to the preceding exercise here as your starting point.
ggplot()
```

::: {#plot_facet-hint}
**Hint:** You must use the vars() function to identify the variable that
is used to create facets.
:::

```{r plot_facet-solution}
# Adaptation: word of mouth as facets instead of shape.
ggplot(
  data = consumers,
  mapping = 
    aes(x = ad_expo,
        y =brand_aw)
  )+
  geom_point(
    mapping = aes(color = gender),
    size = 4
    ) +
  geom_smooth(
    mapping = 
    aes(color = gender),
    method = "lm",
    se = FALSE
    ) +
  labs(
    title = "Does brand awareness depend on exposure, \nword-of-mouth, and gender?",
    x = "Exposure to the campaign",
    y = "Brand awareness"
    ) +
  facet_wrap(~wom)
```

```{r plot_facet-check}
gradethis::grade_code(
  correct = "You can also use facet_wrap(vars(wom))."
)
```

### Adding labels to datapoints

We may want to know who is the outlier in the plot, having an
exceptionally low exposure score. Use `geom_text()` to add the
participants' first names to the plot, as shown below. Carefully inspect
the arguments for this geom.

```{r plot_label_show, echo=FALSE, warning=FALSE, message=FALSE}
# Adaptation: add firstname as label.
ggplot2::ggplot(
  data = consumers,
  mapping = 
    aes(x = ad_expo,
        y = brand_aw)
  )+
  geom_point(
    mapping = aes(shape = wom, color = gender),
    size = 4
    ) +
  geom_smooth(
    mapping = 
    aes(color = gender),
    method = "lm",
    se = FALSE,
    ) +
  labs(
    title = "Does brand awareness depend on exposure, \nword-of-mouth, and gender?",
    x = "Exposure to the campaign",
    y = "Brand awareness"
    ) +
  geom_text(
    aes(label = firstname),
    nudge_y = 0.4,
    check_overlap = TRUE
    )
```

::: question
Create this graph yourself.
:::

```{r plot_label, exercise=TRUE, exercise.lines = 20, warning=FALSE, message=FALSE}
ggplot(
  data = consumers,
  mapping = 
    aes(x = ad_expo,
        y = brand_aw)
  )+
  geom_point(
    mapping = aes(shape = wom, color = gender),
    size = 4
    ) +
  geom_smooth(
    mapping = 
    aes(color = gender),
    method = "lm",
    se = FALSE,
    ) +
  labs(
    title = "Does brand awareness depend on exposure, \nword-of-mouth, and gender?",
    x = "Exposure to the campaign",
    y = "Brand awareness"
    ) +
  geom_text()
```

::: {#plot_label-hint}
**Hint:** Use help on this geom to get more information about the
arguments of this geom.
:::

```{r plot_label-solution}
# Adaptation: swap color and shape, line color as aesthetic.
ggplot(
  data = consumers,
  mapping = 
    aes(x = ad_expo,
        y = brand_aw)
  )+
  geom_point(
    mapping = aes(shape = wom, color = gender),
    size = 4
    ) +
  geom_smooth(
    mapping = 
    aes(color = gender),
    method = "lm",
    se = FALSE,
    ) +
  labs(
    title = "Does brand awareness depend on exposure, \nword-of-mouth, and gender?",
    x = "Exposure to the campaign",
    y = "Brand awareness"
    ) +
  geom_text(
    aes(label = firstname),
    nudge_y = 0.4,
    check_overlap = TRUE
    )
```

```{r plot_label-check}
gradethis::grade_code(
  correct = "The example plot uses `geom_text()` because the boxes created by `geom_label()` obscure the data."
  )
```

### Adding primitives

Actually, we can draw all kinds of shapes on the plot, for example, an
arrow drawing attention to the extremely low exposure score.

::: question
Use `geom_segment` to create this plot.
:::

```{r plot_primitives_show, echo=FALSE, warning=FALSE, message=FALSE}
# Adaptation: Add arrow pointing to extreme value.
ggplot2::ggplot(
  data = consumers,
  mapping = 
    aes(x = ad_expo,
        y = brand_aw)
  )+
  geom_point(
    mapping = aes(shape = wom, color = gender),
    size = 4
    ) +
  geom_smooth(
    mapping = 
    aes(color = gender),
    method = "lm",
    se = FALSE
    ) +
  labs(
    title = "Does brand awareness depend on exposure, \nword-of-mouth, and gender?",
    x = "Exposure to the campaign",
    y = "Brand awareness"
    ) +
  geom_segment(
    x = 1, xend = 1, y = 4, yend = 2,
    arrow = arrow(
      type = "closed"
    )
  )
```

```{r plot_primitives, exercise=TRUE, exercise.lines = 20}
ggplot(
  data = consumers,
  mapping = 
    aes(x = ad_expo,
        y = brand_aw)
  )+
  geom_point(
    mapping = aes(shape = wom, color = gender),
    size = 4
    ) +
  geom_smooth(
    mapping = 
    aes(color = gender),
    method = "lm",
    se = FALSE
    ) +
  labs(
    title = "Does brand awareness depend on exposure, \nword-of-mouth, and gender?",
    x = "Exposure to the campaign",
    y = "Brand awareness"
    ) +
  geom_segment()
```

::: {#plot_primitives-hint}
**Hint:** Pay attention to the `arrow` argument. It uses a function! Use
help on the geom and on this function.
:::

```{r plot_primitives-solution}
# Adaptation: Add arrow pointing to extreme value.
ggplot(
  data = consumers,
  mapping = 
    aes(x = ad_expo,
        y = brand_aw)
  )+
  geom_point(
    mapping = aes(shape = wom, color = gender),
    size = 4
    ) +
  geom_smooth(
    mapping = 
    aes(color = gender),
    method = "lm",
    se = FALSE
    ) +
  labs(
    title = "Does brand awareness depend on exposure, \nword-of-mouth, and gender?",
    x = "Exposure to the campaign",
    y = "Brand awareness"
    ) +
  geom_segment(
    x = 1, xend = 1, y = 4, yend = 2,
    arrow = arrow(
      type = "closed"
    )
  )
```

```{r plot_primitives-check}
gradethis::grade_code(
  correct = "And yes, you can change the color and size of the arrow just like you can change them for points and other geoms."
)
```

### Adding summaries to plots

An important goal when visualizing data is to visually emphasize important
trends in the data. One way to do this is to display summary information
together with all data points. 

To do this, you first need to store summary information about a data set in a data object. 
Every geom can have its own data argument, so specify the data
object with summaries in a geom that visualizes the summary information.

The below graph shows an example based on our (fake) consumer data. It shows the average brand awareness for male and female participants (with jitter) who have high (above median) or low (below median) exposure to the ads as well as their average brand awareness.

```{r prepSumPlot, echo=FALSE}
consumers<-consumers |> 
  mutate(ad_expo2=case_when(ad_expo<median(ad_expo)~"low exposure",ad_expo>=median(ad_expo)~"high exposure"))
```

```{r summaryPlot, exercise = TRUE, exercise.eval = TRUE}
#Calculate and save 
#brand awareness by groups. 
#group by gender and ad exposure groups (above/below median)
averages <- consumers |> group_by(gender,ad_expo2) |> 
  #calculate mean, ignori missing values
  summarise(mean_brand_aw = mean(brand_aw, na.rm = TRUE))
#Create plot
ggplot() + 
  #add individual brand awareness score with jitter 
  geom_jitter(data= consumers, aes(x = ad_expo2, y = brand_aw, color=gender)) +
#add mean scores as crosses
  geom_point(data = averages, aes(x = ad_expo2, y = mean_brand_aw, color = gender), shape='cross',size = 4) +
  #add lines to link the means: note the `group` argument 
  geom_line(data = averages,
aes(x = ad_expo2, y = mean_brand_aw, group = gender, color = gender))
```

###

If we want to visually emphasize some parts of the graph over others, we can further tweak the different
visual elements, for instance by adjusting colors
and sizes. 

::: question
Add transparency to `geom_jitter` to visually emphasize the mean scores. 
:::

```{r summaryPlot2, exercise = TRUE, exercise.eval = TRUE}
ggplot() + 
  geom_jitter(data= consumers, aes(x = ad_expo2, y = brand_aw, color=gender), alpha=0.3) +
  geom_point(data = averages, aes(x = ad_expo2, y = mean_brand_aw, color = gender), shape='cross',size = 4) +
  geom_line(data = averages,
aes(x = ad_expo2, y = mean_brand_aw, group = gender, color = gender))
```

Play around with this plot. For example, use geom_text() to add the
values of the group means to the plot or change the theme to make the background less prominent.

## Debugging

Your friend tried to create a non-stacked bar chart showing the
proportion of females in the consumers data set who heard by word of
mouth against the proportion who did not, as well as the proportion of
males who heard and who did not hear by word of mouth.

::: question
Your friend did not manage to get the code to work. Can you do that?
:::

```{r debug, exercise = TRUE, eval=FALSE}
ggplot(data consumers) +
  geom_bar(
    mapping = aes(
      x = wom, 
      color = gender,
      position = "dodge"
  )
```

<!-- To hide the solution, use a textual hint. -->

::: {#debug-hint}
**Hints:**

-   Pay attention to the error messages when you run the code.
-   For proportions, you have to use a computed variable: check out the
    help on `geom_bar()`.
-   You may want to check the *R for Data Science* book: bar charts are
    tricky.
:::

```{r debug-solution}
ggplot(data = consumers) +
  geom_bar(
    mapping = aes(
      x = wom, 
      y = ..prop..,
      group = gender,
      fill = gender
    ),
    position = "dodge"
  )
```

```{r debug-check}
gradethis::grade_code(
  correct = "To get proportions, you must indeed specify both the y argument (indicating that you want proportions) and the group argument (specifying which total to use for caluclating proportions).", 
  incorrect = "If you don't see the problems, build up the graph from zero, step by step."
  )
```

::: tip
**Programming Tips**

-   When you program something complicated, build it in steps.
-   Check intermediate results regularly by running the code and
    inspecting the results.
:::


## Evaluating a Plot

Once you master the `ggplot2` package and other packages for creating
plots (see the **Fancy stuff** part of this tutorial), you should start
thinking about what you want to convey with a plot and whether the plot
conveys your message in a clear and attractive way.

We will use the following criteria in this course:

1.  The data visualization is sufficiently complex to tell a story. For
    example, it presents both a pattern (as a first impression) and
    deviations from this pattern (inviting reflection about the
    pattern). The deviations, however, should not be overwhelming
    because of too much information.

2.  The data visualization is self-explanatory. It should be
    comprehensible if it is presented by itself. For example, use
    informative, readable labels.

3.  The data visualization gives an accurate, not a biased view of the
    data. For example, sizes should accurately reflect quantity.

4.  The data visualization uses graphic features (size, font type,
    colors, line styles) in such a way that the main parts are stressed
    and unimportant parts remain visually in the background. Justify
    your choices with comments in the code.

If you see a plot (or any other graphic):

-   Have a close look at a plot. What does it tell you?
-   When the plot disappears, try to formulate your first impression of
    the plot, both in terms of the information it gives (What stands out
    in the plot?) and the (dis)pleasure it gives you.
-   When the plot reappears, inspect it more closely. Where do your eyes
    go and which information do you collect during the process of
    looking at the plot.

::: question
Practice with the plots below. Use the checkboxes to show plots, which
will first be shown for a few seconds and, a bit later, for a longer
time.
:::

```{r, echo=FALSE}
fluidPage(
  fluidRow(
    column(2,
      radioButtons("radio", label = h3("Select a plot"),
    choices = list("Plot 1" = 1, "Plot 2" = 2, "Plot 3" = 3, "Plot 4" = 4, "Plot 5" = 5, "Plot 6" = 6), 
    selected = 1),
      sliderInput("slider", label = "", min = 0, max = 20, value = 0,
                  step = 1, ticks = FALSE, 
                  animate = TRUE
      )
    ),
    column(10,
      plotOutput("regPlot")
    )
  )  
)
```

```{r, context="server", warning=FALSE, message=FALSE}
output$regPlot <- renderPlot({
  g <- ggplot(
    data = consumers,
    aes(x = ad_expo, y = brand_aw)
    ) +
    scale_x_continuous(name = "Exposure", breaks = 1:10, limits = c(1, 10)) +
    scale_y_continuous(name = "Brand awareness", breaks = 1:10, limits = c(1, 10)) +
    theme_bw( base_size = 14 )
  
  ## Create plot versions
  if (input$radio == 1) { 
    # just regression line and confidence interval
    g <- g + geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "black")
    }
  else if (input$radio == 2) {
    # point size reflects brand_aw
    g <- g + 
      geom_point(aes(size = brand_aw), color = "grey") +
      geom_smooth(method = "lm", formula = y ~ x, se = FALSE)
    }
  else if (input$radio == 3) {
    # outlier not visible (and regression line without outlier)
    g <- g + 
      geom_point(size = 2, color = "grey") +
      geom_smooth(data = consumers[consumers$ad_expo > 1,], 
        mapping = aes(x = ad_expo, y = brand_aw),
        method = "lm", formula = y ~ x, se = FALSE) +
      scale_x_continuous(limits = c(4, 10))
    }
  else if (input$radio == 4) {
    # grey total regression line, red regression line without outlier
    g <- g + 
      geom_point(size = 2, aes(color = ifelse(ad_expo > 1, "grey", "red")), show.legend = FALSE) +
      geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "grey", size = 2) +
      geom_smooth(
        data = consumers[consumers$ad_expo > 1,], 
        mapping = aes(x = ad_expo, y = brand_aw),
        method = "lm", formula = y ~ x, se = FALSE, color = "red"
        ) +
      geom_text(x = 3.9, y = 4.3, label = "without outlier", size = 3, color = "red", alpha = 0.6) +
      scale_color_manual(values = c("red", "grey"))
    }
  else if (input$radio == 5) {
    # regression line per gender, with additional regression line for males without outlier
    g <- g + 
      geom_point(aes(color = gender), size = 2) +
      geom_smooth(method = "lm", formula = y ~ x, se = FALSE, aes(color = gender)) +
      geom_smooth(
        data = consumers[consumers$ad_expo > 1 & consumers$gender == "male",], 
        mapping = aes(x = ad_expo, y = brand_aw),
        method = "lm", formula = y ~ x, se = FALSE, color = "blue", linetype = "dashed"
        ) +
      geom_text(x = 4.1, y = 3.4, label = "without outlier", size = 3, color = "blue", alpha = 0.6)
    }
  else if (input$radio == 6) {
    # density conours and regression lines per gender
    g <- g + 
      geom_density2d(aes(color = gender),  adjust = 2) +
      geom_smooth(method = "lm", formula = y ~ x, se = FALSE, aes(color = gender))
    }
  
  ## Show plot, depending on slider value
  if (input$slider == 0) {
    #show press button text
    ggplot(data = consumers) + geom_text(x = 0.5, y = 0.5, label = "Press the little play button", size = 12) 
    } 
  else if ((input$slider > 0 & input$slider < 4) | (input$slider > 10)) {
    # ask user to reset the slider 
    if (input$slider == 20) {
      if (input$radio == 3 ) {hor_pos = 7}
      else {hor_pos = 5.5}
      g <- g + geom_text(x = hor_pos, y = 8.5, label = "Reset the slider to 0\nbefore you watch another plot.", size = 8)
      }
    #show plot
    g
    } 
  else {
    # What did you see?
    ggplot(data = consumers) + geom_text(x = 0.5, y = 0.5, label = "What did you see?", size = 18)
    }
})
```

## Additional Resources

Book on ggplot2:

-   Hadley Wickham, ggplot2. Elegant Graphics for Data Analysis.
    Springer. <https://ggplot2-book.org/>

Interactive training:

-   <https://www.datacamp.com/courses/data-visualization-with-ggplot2-1>

## Fancy Stuff

### `gganimate` package

If you want to animate your `ggplot` plot, the package
[`gganimate`](https://gganimate.com/index.html) provides you with tools
to create an animated GIF (with the `gifski` package) or a video (with
the `av` package).

The code below creates an animated gif using
[Gapminder](https://www.gapminder.org/) data on life expectancy, GDP per
capita, and population size by country.

::: question
Play around with the settings that are specific to `animate()`.
:::

Note that it takes quite some time to generate the animation.

```{r gganimate1, exercise=TRUE, exercise.lines = 32, exercise.timelimit = 120, eval=FALSE}
# Install the following packages if they haven't been installed.
library(gganimate)
library(gifski)
library(gapminder) #data used

# This code creates an animated ggplot
g <- ggplot(gapminder::gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  ease_aes('linear') +
  shadow_wake(
    wake_length = 0.2
  )

# And here, we show the animated plot.
gganimate::animate(
  plot = g, #gganimate plot to be shown
  nframes = 78, #1 frame for each year from 1952 to 2007 plus 2x11 additional frames for start and end 
  renderer = gifski_renderer( #save as animated GIF
    file = "gapminder.gif",
    loop = TRUE
    ),
  start_pause = 12, #first frame shows 12 times
  end_pause = 12, #last frame shows 36 times
  rewind = FALSE #roll back to the start
)
```

### 

If you would have a look at the gapminder data (e.g., with
`View(gapminder)`), you would see that the data are available for 1952,
1957, 1962, 1967, and so on. The `gganimate()` animation creates frames
for the years in between. In a way, the data for the in-between-years
are fabricated; they may give a wrong view of reality.

::: tip
**Programming Tip**

-   Carefully inspect the results created by packages that do a lot of
    work for you. They may do things that you do not want.
:::

Movement is fascinating but it can also be frustrating if the user
cannot pause or determine the speed of the animation. *Evaluate the pros
and cons of animations critically.*

### `ggplotly()` in the `plotly` package

The `plotly` library is designed for creating interactive graphics. It
has its own language for creating graphs but for the `ggplot` user, it
provides the 'ggplotly()' function to change a `ggplot` plot into an
interactive `plotly` plot.

The `plotly` library offers the option to zoom, select items in the
graph, and see additional information about the items in the graph.

::: question
Run the code to see the graph.
:::

If you carefully position your cursor on a dot, the respondent's first
name will pop up (works better if you select the *Compare data on hover*
option).

```{r plotly1, exercise=TRUE, exercise.lines = 26}
#these packages have been installed by UsingRTutorials
library(plotly) 
library(gapminder) #contains the data used here

#Step 1: create a ggplot
g <- ggplot(
  data = consumers,
  aes(x = ad_expo, y = brand_aw)
  ) +
    geom_point(aes(color = gender), size = 2) +
    geom_smooth(method = "lm", formula = y ~ x, se = FALSE, aes(color = gender)) +
    geom_smooth(
      data = consumers[consumers$ad_expo > 1 & consumers$gender == "male",], 
      mapping = aes(x = ad_expo, y = brand_aw),
      method = "lm", formula = y ~ x, se = FALSE, color = "blue", linetype = "dashed"
      ) +
    geom_text(
      aes(label = firstname),
      alpha = 0 #trick: make labels invisible (transparent)
      ) +
    scale_x_continuous(name = "Exposure", breaks = 1:10, limits = c(1, 10)) +
    scale_y_continuous(name = "Brand awareness", breaks = 1:10, limits = c(1, 10)) +
    theme_bw()

#Step 2: Turn the ggplot into a plotly plot and use plotly options.
ggplotly(g, tooltip = c("text"), dynamicTicks = TRUE)
```

### 

There are (limited) options for animation, as demonstrated in the plot
below. Run the code to see the plot.

::: question
Change the code, so the pop ups only show the country name.
:::

```{r plotly2, exercise=TRUE}
#Step 1: create a ggplot and use a variable to define the frames of the animation
p <- ggplot(gapminder, aes(gdpPercap, lifeExp, color = continent)) +
  geom_point(aes(size = pop, frame = year, ids = country)) +
  scale_x_log10() #this changes the scale to a log scale, so very large differences are compressed

#Step 2: Turn the ggplot into a plotly plot.
ggplotly(p)
```

::: {#plotly2-hint}
**Hint:** Have a look at the help for the `ggplotly()` function.
:::

```{r plotly2-solution}
p <- ggplot(gapminder, aes(gdpPercap, lifeExp, color = continent)) +
  geom_point(aes(size = pop, frame = year, ids = country)) +
  scale_x_log10() #this changes the scale to a log scale, so very large differences are compressed

ggplotly(p, tooltip = "country")
```

```{r plotly2-check}
gradethis::grade_code()
```

### Networks

Many-to-many relations can be visualized as networks. Friendship
nominations among students can be depicted with nodes representing
students and arrows showing friendship nominations.

The `StudentWaveFriend` tibble contains all the information that we need
to draw a friendship network for the Glasgow data which we worked with in previous sessions:

-   Variable `student` in this tibble represents the student nominating
    a friend.
-   Variable `friend` represents the student nominated.

A line in a network is just a pair of nodes!

There are many R packages for network visualization and network analysis
or modelling. [Network visualization with
R](https://kateto.net/network-visualization) offers a tutorial on some R
packages for network visualization. The examples below use the `igraph`
package.

::: question
Use the tutorial referenced above and the `igraph` documentation to play
around with the friendship network in wave1.
:::

```{r network1-setup}
#preparing data tibbles for network
Student <- GlasgowFriends |> select(student, neighbourhood, schooldist, hoodname, age, sex, smoking_at_home, smoking_parents, smoking_siblings) |> distinct()
StudentWave <- GlasgowFriends |> select(student, wave, alcohol, cannabis, money, romantic, friend_1, friend_2, friend_3, friend_4, friend_5, friend_6 ) |> distinct()
StudentWaveFriend <- StudentWave |> 
  gather(key = "friendnumber", value = "friend", friend_1:friend_6, na.rm = TRUE) |>
  select(student, wave, friend)
```

```{r network1, exercise = TRUE, exercise.lines = 20, exercise.eval = TRUE, fig.asp=0.5, out.width="100%", fig.cap="Figure: Friendship nominations at the first wave. Mutual nominations are black, unilateral nominations are orange."}
# Install the igraph package (once).
#install.packages("igraph")
# Load the igraph package.
library(igraph)
# Retrieve the friendship nominations in the first wave and ensure that the
# first two columns are the sender and receiver of the nomination.
arrows_wave1 <- StudentWaveFriend |>
  filter(wave == "t1") |>
  select(student, friend)
#add color variable for sexes
nodes_wave1 <- Student |>
  mutate(sex_color = ifelse(sex == "boy", "steelblue1", "tomato1"))
# Create an igraph object: a list of nodes (=vertices) combined with a list of lines/arrows.
net_wave1 <- igraph::graph_from_data_frame(d=arrows_wave1, vertices=nodes_wave1, directed=TRUE)

# Create a layout: positioning of the nodes.
set.seed(35546)
layout_1 <- igraph::layout_with_fr(net_wave1)
# Spread out the network horizontally.
layout_1 <- igraph::norm_coords(layout_1, ymin=-1, ymax=1, xmin=-2, xmax=2)
# Plot the network.
par(mar = c(0, 0, 0, 0))
plot(net_wave1, 
     vertex.label = NA, #hide node labels
     vertex.size = 8, #vertex size
     vertex.color = V(net_wave1)$sex_color,
     edge.arrow.size=.4, # arrowhead size
     #give reciprocal arrows (edges) a different colour than unilateral arrows:
     edge.color = ifelse(which_mutual(net_wave1), "gray40", "tan1"),
     rescale = FALSE, #don't rescale to square network
     layout = layout_1 #use the previously created node coordinates
     )
legend("topleft", c("Boys", "Girls"), pch=21, col="#777777", pt.bg=c("steelblue1", "tomato1"), pt.cex=2, cex=.8, bty="n", ncol=1)
```

### A geographic network

With `ggplot()` and the `map::` package, we can plot geographical
networks.

For example, the flights to/from NY City in the `nycflights13::`
(example data in the book).

```{r nycflightsmap, exercise = TRUE, exercise.eval = TRUE}
# Load the nycflights13 package.
library(nycflights13)
# Connect the flights (origin & destination) and airports tables (airport coordinates).
flights |> #use the flights tibble in the nycflights13 package
  #add coordinates to airport where flight originates
  inner_join(select(airports, origin = faa, origin_lat = lat, origin_lon = lon), 
             by = "origin") |>
  #add coordinates to destination airport
  inner_join(select(airports, dest = faa, dest_lat = lat, dest_lon = lon), by = "dest"
  ) |>
  # select the first 100 flights for illustrative purposes
  slice(1:100) |>
  # pipe the data into ggplot()
  ggplot(aes(
    x = origin_lon, xend = dest_lon,
    y = origin_lat, yend = dest_lat
    )) +
    borders("state") +
    geom_segment(arrow = arrow(length = unit(0.1, "cm"))) +
    coord_quickmap() +
    labs(y = "Latitude", x = "Longitude")
```

## What's next

- We are in the middle of Sprint 4. Keep working on your Sprint Backlog. 
- Good luck with the group part deadline!
- You now have important information relevant to your individual Data Project. 
- Prepare for next session's individual consultations by starting to work on your individual part. Come with questions!
