---
title: 'Session 5: Programming'
output:
  learnr::tutorial:
    progressive: yes
    allow_skip: yes
runtime: shiny_prerendered
description: |
  Practice with creating functions and some basic programming tricks. As of this session, there is more time for working on your Data Project.
---

```{r setup, include=FALSE}
# Ensure that libraries are loaded.
library(tidyverse)
library(learnr)
library(gradethis)
library(knitr)
library(kableExtra)
# New packages (must be installed before taking this tutorial)

tutorial_options(exercise.timelimit = 60, exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r, context="data", include=FALSE}
# Ensure that the data is loaded for the remainder of this tutorial.
flights4 <- UsingRTutorials::flights4
# Store the results of a t test, so the htest object is available in the tutorial.
result_t <- flights4 |> t.test(arr_delay0 ~ origin, data = ., na.action = "na.omit")
# Ensure that the final function report_ttest() is available in the tutorial.
report_ttest <- function(result_t) { 
  # if (is.null(result_t) || class(result_t) != "htest") { 
  #   return("#### Input is not a result of t.test()! ######") 
  # } else { 
      paste0("*t* (", format(round(result_t$parameter, digits = 2), nsmall = 2), ") = ",
             format(round(result_t$statistic, digits = 2), nsmall = 2), ", *p* ", 
             ifelse(result_t$p.value >= 0.0005, 
                    paste0("= ", format(round(result_t$p.value, digits = 3), nsmall = 3)),
                    "< 0.001"), 
             ", 95%CI[", format(round(result_t$conf.int[1], digits = 2), nsmall = 2), 
             ", ", format(round(result_t$conf.int[2], digits = 2), nsmall = 2), "]") 
  # } 
}
```

<!-- Define programming tip style -->
<style>
.tip {
  background-color: #f5f5f5;
}
</style>

<!-- Define question style -->
<style>
.question {
  color: #5A9DDB;
}
</style>

<!-- Define emphasis style -->
<style>
.emphasis {
  color: #e8301b;
}
</style>

## Overview

First 1.5 hours: Course content

-   Data Tidying: Expert meeting
-   Piping
-   Lists
-   Functions
-   Control Flow: Conditions

Second 1.5 hours: Data project

-   Continue work on Sprint 3.
-   Ask questions.

## Piping

### 

<div class="emphasis" >
Pipe more: Cut out intermediary results. 
</div>

Let us quickly rehearse piping.

<div class="question" >
The code below creates and shows a plot. Replace it by one pipe that is short as possible.
</div>

```{r pipeIt, exercise = TRUE, exercise.lines = 7}
# Show a plot of the logarithm of arrival delays for flights to 
# Atlanta, Boston, or Buffalo in January.
flights4_january <- filter(flights4, month == 1)
flights4_jan_atlantabostonbuffalo <- filter(flights4_january, dest %in% c("ATL", "BOS", "BUF"))
flights4_jan_ATLBOSBUF_logdelay <- mutate(flights4_jan_atlantabostonbuffalo, log_arr_delay0 = log(arr_delay0 + 1))
ggplot(flights4_jan_ATLBOSBUF_logdelay, aes(x = log_arr_delay0)) +
  geom_area(stat = "count")
```

<!-- To hide the solution, use a textual hint. -->
<div id="pipeIt-hint">
__Hint:__ What do we do with the data argument in a pipe?
</div>

```{r pipeIt-solution}
flights4 |> filter(month == 1 & dest %in% c("ATL", "BOS", "BUF")) |> mutate(log_arr_delay0 = log(arr_delay0 + 1)) |> ggplot(aes(x = log_arr_delay0)) + geom_area(stat = "count")
```

```{r pipeIt-check}
gradethis::grade_code(
  correct = "The (input) data argument disappears in a pipe if it is the first argument. And you correctly combined the two filter steps.", 
  incorrect = "Don't save the plot as a data object, send it to the screen. Perhaps you should join the two filter functions."
  )
```

###

The pipe:

-   much more concise;
-   no intermediary data objects that clutter your environment;
-   no intermediary data objects must be specified as argument;
-   no risk of inadvertently overwriting an intermediary data object.

### `.` in a pipe  

Piping:

-   First argument of function must be a tibble/data frame. 
-   All `tidyverse` functions have this characteristic. 
-   Base R functions do not necessarily have this characteristic.

<div class="question" >
Test if average arrival delay (`arr_delay0`) differs between the two airports of origin (`origin`) included in the `flights4` tibble. Pipe this tibble into the `t.test()` function (of the `stats` package).
</div>

```{r pipedot, exercise = TRUE}

```

```{r pipedot-hint-1}
# Check help on the `t.test()` function.
```

```{r pipedot-hint-2}
# Use a dot to specify the piped-in tibble as input data.
flights4 |> t.test(arr_delay0 ~ origin, )
```

```{r pipedot-solution, exercise.reveal_solution = FALSE}
flights4 |> t.test(arr_delay0 ~ origin, data = .)
```

```{r pipedot-check}
gradethis::grade_code(
  correct = "The `data` argument is not the first argument, so it must be specified and the input data tibble must be represented by a dot. As you have done!", 
  incorrect = "Did you use the data argument in the t.test() function?"
  )
```

###

In a pipe, `.` represents the input object. Use it if:  

- the data frame is not the first argument of a function (see the preceding exercise),  
- you don't use a function (see the example below).  

```{r pipedot2, exercise = TRUE, exercise.lines = 4, exercise.eval = TRUE}
# Get the value of the t statistic of an independent samples t test.
flights4 |> 
  t.test(arr_delay0 ~ origin, data = .) |>
  .$statistic #not a function
```

We will see soon where `$statistic` comes from.  

## Lists

### 

-   Vector: Collection of one type of elements.
-   List: Collection of (possibly) multiple types of elements.

A `List` may store anything (your perfect cupboard?)

Examples:  

1. Data frame (or tibble): list of (logic/num/char) vectors (variables).  
    Restriction: All vectors have the same length.  
2. Statistical results: list with coefficients, etc.  
3. Nested data: More than one value per variable per case.  

### Extracting statistical results  

A _t_ test yields a results object, which is a list.  

```{r storeresults, exercise = TRUE, exercise.lines = 4}
# Store the results of a t test.
result_t <- flights4 |> 
  t.test(arr_delay0 ~ origin, data = .)
result_t #Default print method for these results.
```

- Results of t test are saved as data object `result_t` here, so we can inspect the results object in the Environment (if you run this code in RStudio). 

### 

Function `str()` shows the contents of a list.

<div class="question" >
Try to recognize the contents of this list.
</div>

```{r str, exercise=TRUE, exercise.eval = TRUE}
str(result_t)
```

-   List elements are different: one number, two numbers, character string.  
-   "names" attribute: the name(s) of the list element value(s).
-   other attribute: other description of the list element value, e.g., the confidence level of the confidence interval.

Note:

-   _t_ test results are stored as object of class: `htest` (bottom of the output);
-   Other R functions (e.g., `print()`) know what to do with the list.

### Extracting an element from a list

We get elements from a list:

-   Appending the element name between `[[]]` to the list name;
-   Appending the element name with `$` to the list name.

<div class="question" >
Extract the confidence interval from the t test results data object (`result_t`) with `[[]]` and show (print) it (instead of saving it as a new data object). 
</div>

```{r extractbracket, exercise = TRUE}

```

<!-- To hide the solution, use a textual hint. -->
<div id="extractbracket-hint">
__Hint:__ The element name must be in quotation marks. See the book section on _Recursive Vectors (Lists)_.
</div>

```{r extractbracket-solution}
result_t[["conf.int"]]
```

```{r extractbracket-check}
gradethis::grade_code(
  correct = "", 
  incorrect = "Perhaps you used the element number in the list instead of the element name to extract the confidence level. That is OK."
  )
```

<div class="question" >
Do the same thing but now use `$`.
</div>

```{r extractdollar, exercise = TRUE}

```

<!-- To hide the solution, use a textual hint. -->
<div id="extractdollar-hint">
__Hint:__ See the book section on _Recursive Vectors (Lists)_.
</div>

```{r extractdollar-solution}
result_t$conf.int
```

```{r extractdollar-check}
gradethis::grade_code(
  correct = "`$conf.int` is shorthand for `[['conf.int']]`.", 
  incorrect = ""
  )
```

###

Let's go one level down in the list of t test results: 

-   Extract the value of an attribute with the `attribute()` function.

```{r}
str(result_t[4])
```

<div class="question" >
Show the confidence level (as a number) using only dollar signs (`$`), not square brackets (`[[]]`). Show the result on the screen. 
</div>

```{r extractlevel, exercise = TRUE}

```

```{r extractlevel-hint-1}
# The confidence level is an attribute (`attr` in the structure overview presented above), so use the `attributes()` function.
attributes()
```

```{r extractlevel-hint-2}
# Pull out the confidence interval from the results. Complete this code.
attributes(result_t)
```

```{r extractlevel-hint-3}
# Have a look at the output of the attributes() function. How can you get the number of the confidence level?
attributes(result_t$conf.int)
```

```{r extractlevel-solution, exercise.reveal_solution = FALSE}
attributes(result_t$conf.int)$conf.level
```

```{r extractlevel-check}
gradethis::grade_code(
  correct = "We can use the dolar sign directly after the attributes() function. Isn't that nice?", 
  incorrect = "If you just get the number 0.95, you are fine. Probably, you used [[]] instead of $. If you get more than just the number, you have to go down one level in the list that `attributes()` generates."
  )
```

###

Instead of the attribute value, we can get the attribute name with the `names()` function.

<div class="question" >
Extract the label "conf.level" from the list of t test results.
</div>

```{r extractlabel, exercise = TRUE}

```

```{r extractlabel-hint-1}
# If you want the name of an attribute, apply the `names()` function to an attribute that you extract with attributes().
names(attributes( ))
```

```{r extractlabel-solution, exercise.reveal_solution = FALSE}
names(attributes(result_t$conf.int))
```

```{r extractlabel-check}
gradethis::grade_code(
  correct = "", 
  incorrect = "Did you use [[]] instead of $?"
  )
```

###

Let us practice some more.

<div class="question" >
Get the name and value (rounded to one decimal place) of the mean of the second group in the results of the _t_ test (`result_t`).
</div>

```{r paste0, exercise = TRUE}

```

```{r paste0-hint-1}
# Build your code in steps. First, find the group means in the results.
str(result_t)
```

```{r paste0-hint-2}
# Second, pull the mean from the results data object. You know how to get only the second mean (complete the code below).
result_t$estimate
```

```{r paste0-hint-3}
# Third, round the mean with the round() function (use Help). Complete the code below.
round(result_t$estimate[[2]] )
```

```{r paste0-hint-4}
# Fourth, pull the name of the second group from the results data object. Complete the code below.
names(result_t$estimate)[2]
```

```{r paste0-solution, exercise.reveal_solution = FALSE}
names(result_t$estimate)[2]
round(result_t$estimate[[2]], digits = 1)
```

```{r paste0-check}
gradethis::grade_code(
  correct = "", 
  incorrect = "Check out the hints to this exercise. And mind the blanks in the resulting sentence."
  )
```

###

<div class="tip" >
__Programming Tips__

- Attribute names and attribute values are nested within a list element.  
- `attributes()` creates a list; select from this list with $ or [[]].
- 'Drill' into lower levels step by step: 
    + `attributes(result_t)` returns all first-level names and attributes (as a list);
    + `attributes(result_t$conf.int)` returns all attributes of this element;
    + `attributes(result_t$conf.int)$conf.level` gives the value of the conf.level attribute;  
    + `names(attributes(result_t$conf.int))` returns the name of the conf.int attribute.  
- `result_t$conf.int[1]` yields the lower limit of the confidence interval
</div>

### Extract test results in APA style

The code below extracts the results of the t test in APA style.  

<div class="question" >
What does the `digits` argument do?
</div>

```{r APA1, exercise = TRUE, exercise.eval = TRUE, exercise.lines = 7}
paste0(                       # base R function to concatenate strings
  "*t* (", 
  round(result_t$parameter, digits = 2), ") = ",        # df
  round(result_t$statistic, digits = 2),                # t
  ", *p* = ", round(result_t$p.value, digits = 3),      # p
  ", 95%CI[", round(result_t$conf.int[1], digits = 2),  # 95%CI lower
  ", ", round(result_t$conf.int[2], digits = 2), "]")   # 95%CI upper
```

The stars will turn *t* and *p* into italics in a knitted document.

### Conditions with `ifelse()` 

Improving the reported p value: 

-   either ‘p < 0.001’ 
-   or the exact p value.

<div class="question" >

1.   What does the `digits` argument do?
2.   To which function does it belong?
3.   Explain how 'ifelse()` works.

Experiment with the code to check your answers.
</div>

```{r APA2, exercise = TRUE}
paste0("*t* (", 
  format(round(result_t$parameter, digits = 2), nsmall = 2), ") = ",
  format(round(result_t$statistic, digits = 2), nsmall = 2), 
  ", *p* ", 
  ifelse(result_t$p.value < 0.0005, 
         paste0("= ", format(round(result_t$p.value, digits = 3), nsmall = 2)),
         "< 0.001"),
  ", 95%CI[", format(round(result_t$conf.int[1], digits = 2), nsmall = 2),
  ", ", format(round(result_t$conf.int[2], digits = 2), nsmall = 2), "]")
```

<!-- To hide the solution, use a textual hint. -->
<div id="APA2-hint">
__Hint:__ Use help on functions `format()` and `ifelse()`.
</div>

## Functions

### 

Functions in mathematics: `y = f(x)`.

Functions in R: `y <- f(x)`.

Meaning: Do something to data object `x` -- `f(x)` -- and store result in data object `y`. 

For short: Transform `x` into `y`.  

Left-hand data object (`y`):  
- Does not exist: new data object created. Can be a function!  
- Exists: data object overwritten.  
- Not named: output to screen (console).  

### Creating a function for reporting t tests in APA style 

Step 1: Add `function()`, enclose code within `{` and `}`, and store.  

```{r, eval=FALSE, echo=TRUE}
report_ttest <- function() {
  paste0("*t* (", format(round(result_t$parameter, digits = 2), nsmall = 2), ") = ",
    format(round(result_t$statistic, digits = 2), nsmall = 2), ", *p* ", 
    ifelse(result_t$p.value >= 0.0005, 
      paste0("= ", format(round(result_t$p.value, digits = 3), nsmall = 3)),"< 0.001"),
    ", 95%CI[", format(round(result_t$conf.int[1], digits = 2), nsmall = 2),
    ", ", format(round(result_t$conf.int[2], digits = 2), nsmall = 2), "]")
}
```

`report_ttest` is the name of the new function.  

<div class="tip" >
__Programming Tip__

- Start function creation with code that works.
</div>

### Identify inputs as arguments

Step 2: Specify user input. 

The user must specify the data object containing the results of `t.test()`. 

-   In the current code, this object is `result_t`.
-   Recommended: Change everywhere into an abstract name, e..g, __x__.
-   Add it as an argument name.

<div class="question" >
Specify the results data object as __x__ in the below code.
</div>

```{r funstep2, exercise=TRUE, exercise.lines = 8}
report_ttest <- function() {
  paste0("*t* (", format(round(result_t$parameter, digits = 2), nsmall = 2), ") = ",
    format(round(result_t$statistic, digits = 2), nsmall = 2), ", *p* ", 
    ifelse(result_t$p.value >= 0.0005, 
      paste0("= ", format(round(result_t$p.value, digits = 3), nsmall = 3)),"< 0.001"),
    ", 95%CI[", format(round(result_t$conf.int[1], digits = 2), nsmall = 2),
    ", ", format(round(result_t$conf.int[2], digits = 2), nsmall = 2), "]")
}
```

```{r funstep2-solution, exercise.reveal_solution = FALSE}
report_ttest <- function(x) {
  paste0("*t* (", format(round(x$parameter, digits = 2), nsmall = 2), ") = ",
    format(round(x$statistic, digits = 2), nsmall = 2), ", *p* ", 
    ifelse(x$p.value >= 0.0005, 
      paste0("= ", format(round(x$p.value, digits = 3), nsmall = 3)),"< 0.001"),
    ", 95%CI[", format(round(x$conf.int[1], digits = 2), nsmall = 2),
    ", ", format(round(x$conf.int[2], digits = 2), nsmall = 2), "]")
}
```

```{r funstep2-check}
gradethis::grade_code(
  correct = "", 
  incorrect = "Don't forget to add x as an argument to the function. Replace result_t everywhere by x!"
  )
```

### Check encapsulation

Encapsulation: 

-   The function does not use data objects that happens to be in the current working environment. 
-   If that is the case, your function will not work (as intended) in another working environment. 

Step 3: Ensure that all data object names in the body are:

1. named as arguments (for example: `x`), or  
2. created within the body (none in this example).  

<div class="question" >
What goes wrong and why in the `report_ttest()` function? Can you solve the problem?
</div>

```{r APA4, exercise = TRUE, exercise.lines = 17}
# The function is created here.
report_ttest <- function(x) {
  paste0("*t* (", format(round(x$parameter, digits = 2), nsmall = 2), ") = ",
    format(round(x$statistic, digits = 2), nsmall = 2), ", *p* ", 
    ifelse(x$p.value >= 0.0005, 
      paste0("= ", format(round(x$p.value, digits = 3), nsmall = 3)),"< 0.001"),
    ", 95%CI[", format(round(result_t$conf.int[1], digits = 2), nsmall = 2),
    ", ", format(round(result_t$conf.int[2], digits = 2), nsmall = 2), "]")
}
# Execute another t test: average arrival delay of carriers AA and UA.
result_t2 <- flights4 |> 
  filter(carrier %in% c("AA", "UA")) |>
  t.test(arr_delay0 ~ carrier, data = .)
# New test results.
result_t2
# New test results with function report_ttest().
report_ttest(result_t2)
```

<div id="APA4-hint">
__Hint:__ Carefully compare the results. Note that `result_t` is present in the environment (you can check this with the `ls()` function.
</div>

### 

-   A function assumes that all packages used have been installed by the user. 
-   Include a `library()` command in the function for every package used in the function.
-   Standard packages like `base` and `stats` are automatically loaded.
-   Here, functions `paste0()`, `format()`, and `round()` are used, which are part of the `base` package.

### Dot-Dot-Dot

`...` is a special argument for R functions:

-   Captures any arguments, so they don't yield errors.
-   Most useful to send arguments to a function used inside (i.e. wrapped by) your function.

<div class="question" >
What is the input for the dots in the last line of code and what does the function do with this input?
</div>

```{r APA3dotdotdot, exercise = TRUE, exercise.lines = 11}
report_ttest <- function(x, ...) {
  print(paste(...))
  paste0("*t* (", format(round(x$parameter, digits = 2), nsmall = 2), ") = ",
    format(round(x$statistic, digits = 2), nsmall = 2), ", *p* ", 
    ifelse(x$p.value >= 0.0005, 
      paste0("= ", format(round(x$p.value, digits = 3), nsmall = 3)),"< 0.001"),
    ", 95%CI[", format(round(x$conf.int[1], digits = 2), nsmall = 2),
    ", ", format(round(x$conf.int[2], digits = 2), nsmall = 2), "]")
}
# Use the function.
report_ttest(result_t, "The", "difference", "was", "significant.", sep=' ')
```

### Named arguments and default values

<div class="question" >
1.    Add a named argument `digits =` to the `report_ttest()` function that specifies the number of decimal places for numeric results other than _p_ values.
2.    Set the default number of decimal places to 2 in this function.
3.    And ensure that the requested number of decimal places are used in the function output.
</div>

```{r APA3, exercise = TRUE, exercise.lines = 10}
report_ttest <- function(x) {
  paste0("*t* (", format(round(x$parameter, digits = 2), nsmall = 2), ") = ",
    format(round(x$statistic, digits = 2), nsmall = 2), ", *p* ", 
    ifelse(x$p.value >= 0.0005, 
      paste0("= ", format(round(x$p.value, digits = 3), nsmall = 3)),"< 0.001"),
    ", 95%CI[", format(round(x$conf.int[1], digits = 2), nsmall = 2),
    ", ", format(round(x$conf.int[2], digits = 2), nsmall = 2), "]")
}
# Test the function.
report_ttest(result_t, digits = 5)
```

<!-- To hide the solution, use a textual hint. -->
<div id="APA3-hint">
__Hint:__ The name of the new argument must be used within the function everywhere the number of digits should follow the choice of the user.
</div>

```{r APA3-solution}
report_ttest <- function(x, digits = 2) { paste0("*t* (", format(round(x$parameter, digits = digits), nsmall = digits), ") = ", format(round(x$statistic, digits = digits), nsmall = digits), ", *p* ",  ifelse(x$p.value >= 0.0005,  paste0("= ", format(round(x$p.value, digits = 3), nsmall = 3)),"< 0.001"), ", 95%CI[", format(round(x$conf.int[1], digits = digits), nsmall = digits), ", ", format(round(x$conf.int[2], digits = digits), nsmall = digits), "]") }
report_ttest(result_t, digits = 5)
```

```{r APA3-check}
gradethis::grade_code(
  correct = "", 
  incorrect = "You must use digits instead of the number 2 everywhere. And don't forget to specify the default number of digits in the function's argument."
  )
```

### Embedding output in R Markdown

Function `report_ttest()` is meant to display _t_ test results in APA style to the reader.

This means that the result of the function must be displayed within a sentence in the report.

This is called i

Inline code: 

-   R results displayed within text for report reader. 
-   Calculated during document knitting: always up-to-date.

Example R Markdown text with inline code:

```{r, eval=FALSE, echo=TRUE}
There is a statistically significant difference in average delay between the two airports, `r report_ttest(result_t)`.  
```

The R Markdown text shown above if the document is knitted: 

> There is a statistically significant difference in average delay between the two airports, `r report_ttest(result_t)`.  

Note the italics of _t_ and _p_.

<div class="tip" >
__Programming Tip__

- This is a valuable way to report correct numerical results. If the data or analysis change, the new results will be shown.
</div>

## Control Flow: Conditions

Our function still has flaws: It gives errors or wrong output if the input is not a `t.test()` result (class `htest`).  

```{r setup_resultlm}
#Run a regression.
result_lm <- flights4 |> lm(arr_delay0 ~ origin, data = .)
```

```{r error=TRUE, echo=TRUE}
#Run a regression.
result_lm <- flights4 |> lm(arr_delay0 ~ origin, data = .)
#Use regression results in function report_ttest().
report_ttest(result_lm)
```

### Using `if () {} else {}`

Let's fix it.  

<div class="question" >
Use `if (class(x) != "htest") {} else {}` to print either #### Input is not a result of t.test()! ###### or the APA-style formatted t test result.
</div>

```{r classcheck, exercise = TRUE, exercise.setup = "setup_resultlm", exercise.lines = 14}
report_ttest <- function(x) {
  paste0("*t* (", format(round(x$parameter, digits = 2), nsmall = 2), ") = ",
    format(round(x$statistic, digits = 2), nsmall = 2), ", *p* ", 
    ifelse(x$p.value >= 0.0005, 
      paste0("= ", format(round(x$p.value, digits = 3), nsmall = 3)),"< 0.001"),
    ", 95%CI[", format(round(x$conf.int[1], digits = 2), nsmall = 2),
    ", ", format(round(x$conf.int[2], digits = 2), nsmall = 2), "]")
}
# Applied to t test results.
report_ttest(result_t)
# Applied to regression results.
report_ttest(result_lm)
# Applied to empty object.
report_ttest(NULL) 
```

```{r classcheck-hint-1}
# Check the use of `return()` in a function.
```

```{r classcheck-hint-2}
# The if - else flow control must be at the start of the function, because it makes no sense to pull results from a data object containing something else than a t test result.
report_ttest <- function(result_t) { 
  if (class(result_t) != "htest") { 
    
  } else {
    
  }
}
```

```{r classcheck-hint-3}
# Just return a message if the data object is not of class htest.
report_ttest <- function(result_t) { 
  if (class(result_t) != "htest") { 
    return("")
  } else {
    
  }
}
```

```{r classcheck-hint-4}
# Put the code to create the formatted results in the else{} part.
report_ttest <- function(result_t) { 
  if (class(result_t) != "htest") { 
    return("#### Input is not a result of t.test()! ######")
  } else {
    # add code here
  }
}
```

```{r classcheck-solution, exercise.reveal_solution = FALSE}
report_ttest <- function(x) { if (class(x) != "htest") { return("#### Input is not a result of t.test()! ######") } else { paste0("*t* (", format(round(x$parameter, digits = 2), nsmall = 2), ") = ", format(round(x$statistic, digits = 2), nsmall = 2), ", *p* ",  ifelse(x$p.value >= 0.0005, paste0("= ", format(round(x$p.value, digits = 3), nsmall = 3)),"< 0.001"), ", 95%CI[", format(round(x$conf.int[1], digits = 2), nsmall = 2), ", ", format(round(x$conf.int[2], digits = 2), nsmall = 2), "]") } }
report_ttest(result_t)
report_ttest(result_lm)
report_ttest(NULL) 
```

```{r classcheck-check}
gradethis::grade_code(
  correct = "A function returns the result of the last step in the code or what is marked by `return()`.", 
  incorrect = "If you get the message 'non-numeric argument to mathematical function', input other than t test results is still being treated as if it contains t test results." 
  )
```

### 

A function returns: 

- the last value that it computed, or  
- the output of an explicit `return()` command, which stops further execution of the function.  

<div class="tip" >
__Programming Tip__

- Empty data objects may cause problems. Always include them in a function test.
</div>

### Conditions and logical operators

An _F_ test also yields results as a `htest` class.

How can we ensure that `report_ttest()` does not report _F_ test results?

<div class="question" >
Explain how the code below excludes _F_ test results. What does `||` mean?
</div>

```{r logical, exercise = TRUE, exercise.lines = 17}
report_ttest <- function(x) { 
  if (class(x) != "htest" || x$method == "F test to compare two variances") { 
    return("#### Input is not a result of at.test()! ######") 
  } else { 
      paste0("*t* (", format(round(x$parameter, digits = 2), nsmall = 2), ") = ",
             format(round(x$statistic, digits = 2), nsmall = 2), ", *p* ", 
             ifelse(x$p.value >= 0.0005, 
                    paste0("= ", format(round(x$p.value, digits = 3), nsmall = 3)),
                    "< 0.001"), 
             ", 95%CI[", format(round(x$conf.int[1], digits = 2), nsmall = 2), 
             ", ", format(round(x$conf.int[2], digits = 2), nsmall = 2), "]") 
  } 
}
# Applied to previous t test results.
report_ttest(result_t)
# Applied to F test results.
flights4 |> var.test(arr_delay0 ~ origin, data = .) |> report_ttest()
```

###

In the `report_ttest()` function, we deal with single values instead of vectors:  

- Use double `||` (or) and `&&` (and) instead of `|` and `&`. The latter are for vectorised operations, e.g., in `mutate()`.  
- `==` is vectorised. Unless you're sure that there is one value (as in `class()`), use non-vectorised `identical()`.  

## Fancy Stuff

Sorry, no fancy stuff in this tutorial.

## Data Project Today

-   Continue working on Sprint 3.
-   Ask questions.

It is time to finish tidying your group's project data. 

Use the remaining time to design and create your data visualization. 


