---
title: "Session 5: Review and Q&A"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
    Review content about R basics, visualization, and data tidying.
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
# Ensure that libraries are loaded.
library(tidyverse)
library(learnr)
library(gradethis)
library(knitr)
library(kableExtra)

tutorial_options(exercise.timelimit = 60, exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r, context="data", include=FALSE}
#Load any datasets
```

<!-- Define programming tip style -->

```{=html}
<style>
.tip {
  background-color: #f5f5f5;
}
</style>
```
<!-- Define question style -->

```{=html}
<style>
.question {
  color: #5A9DDB;
}
</style>
```
<!-- Define emphasis style -->

```{=html}
<style>
.emphasis {
  color: #e8301b;
}
</style>
```
## Overview

-   Meet the Experts
-   Review Course Content
-   Group Consultations

## Meet the Experts

Each group has at least one expert per tidy data principle. Each expert
prepared by:

-   Re-visiting the tidy data principle in the book and course
    materials.
-   Applying it to one of your Data Project data sets.
-   Producing 2 PowerPoint slides (use the template in Session 5 module
    on Canvas).

Today, the experts of different groups will meet to discuss their
knowledge and examples.

## Meet the Experts: What to do

Team up with the experts of your topic.

-   Each expert:

    -   Describes the tidy data principle in 1-2 sentences.

    -   Shows one example from their complex dataset to the group of
        experts.

-   Each expert group:

    -   Consolidate your findings in one common power point
        presentation, containing:

        -   1 slide with the best description of the principle the group
            agrees on (WHAT, HOW, WHY).

        -   1 slide PER EXPERT with an example (ONLY INCLUDE GOOD
            EXAMPLES THE GROUP AGREES ON.):

            -   Screenshot of original variable

            -   Screenshot of code to tidy the variable;

            -   Screenshot of end result

## Course Content Review

[ADD SLIDES IF USEFUL]

## Data Project Consultations

Each group will have a consultation with the instructor to discuss their
Data Project. We will have X minutes per group.

| Order | Group        |
|-------|--------------|
| 1     | GROUP NAME 1 |
| 2     | GROUP NAME 2 |
| 3     | GROUP NAME 3 |

## Data Project: What's next

We are in the middle of Sprint 3.

-   Take a moment to update your product and sprint backlogs if needed
    based on the feedback you got today.

-   Keep working on your current sprint backlog.
