---
title: 'Session 14: Your Future with R'
output:
  learnr::tutorial:
    progressive: yes
    allow_skip: yes
runtime: shiny_prerendered
description: |
  Easy work flows and strategies to keep using R in collaborative, open science research.
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
# Ensure that libraries are loaded.
library(tidyverse)
library(learnr)
library(gradethis)
library(haven) #For importing SPSS data files.

tutorial_options(exercise.timelimit = 20, exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r, context="data", include=FALSE}
# Ensure that the data is loaded for the remainder of this tutorial.
glbwarm <- UsingRTutorials::glbwarm
glbwarm_spss <- UsingRTutorials::glbwarm_spss
```

<!-- Define programming tip style -->

```{=html}
<style>
.tip {
  background-color: #f5f5f5;
}
</style>
```
<!-- Define question style -->

```{=html}
<style>
.question {
  color: #5A9DDB;
}
</style>
```
<!-- Define emphasis style -->

```{=html}
<style>
.emphasis {
  color: #e8301b;
}
</style>
```
## Overview

-   Commit to switching

-   Keep learning & teaching

-   Collaborate

-   Practice Open Science

## Commit to Switching

-   Becoming truly proficient in R requires lots of practice.

-   We recommend to fully commit to switching whenever possible to give
    yourself the best chances of succeeding.

-   Do you need to edit, sort, or check something in an Excel sheet?
    Clean data? Make a figure? Make a table with descriptives? Run an
    analysis? –\> Do it in R!

::: question
Think about any ongoing or upcoming data-related projects. What could be
done in R? What not? Why?
:::

### 

### (Reference) Importing SPSS Data

SPSS data files have a complicated setup with variable labels and value
labels.

R data frames or tibbles do not have such labels.

In case you later have SPSS data that you want to analyze in R, here are
**two options** for importing SPSS data.

#### Option 1. Export from SPSS to .csv and import .csv in R.

-   Export data from SPSS to a CSV file with value labels for
    categorical variables:
    -   *File \> Export \> CSV Data* with *Save value labels where
        defined instead of data values*.
-   Use `read_csv()` (as you learned before) to import the CSV file.

::: question
Import the SPSS file `data/glbwarm.csv` and have a look at it.
:::

```{r importCSV, exercise = TRUE, exercise.eval = TRUE}
# The CSV is available in the data directory of this tutorial.
glbwarm <- read_csv("data/glbwarm.csv")
# Inspect the variables.
str(glbwarm)
```

#### Option 2. Import SPSS .sav file directly with tidyverse package `haven`.

The tidyverse package `haven` contains function `read_sav()` (or
`read_spss()`) for importing SPSS (and other software packages) data
files.

::: question
Import the SPSS file `data/glbwarm.sav` and have a look at it.
:::

```{r importSPSS, exercise = TRUE}
# The SPSS system file is available in the data directory of this tutorial.
glbwarm_spss <- haven::read_sav("data/glbwarm.sav")
# Inspect the variables.
str(glbwarm_spss)
```

### 

Imported categorical variables such as *ideology*:

-   Are numerical codes with labels.
-   So, R treats these variables as numerical.
-   If you don't want that, change them into factors.

::: question
Use the `haven::as_factor()` function to add a variable named `sex_fct`
to tibble `glbwarm_spss`.
:::

```{r sex_fct, exercise = TRUE}
# Add sex as a factor to the tibble.
glbwarm_spss <- glbwarm_spss |> 
  mutate(sex_fct = haven::as_factor(sex))
# Inspect the original and new variable.
glbwarm_spss |> count(sex, sex_fct)
```

### 

Passing the entire tibble `glbwarm_spss` to `haven::as_factor()`will
change all labelled variables into factors.

I am not sure this is what you want here. Perhaps you would like to use
the `ideology` variable as a seven-point scale.

## Keep Learning & Teaching

::: emphasis
If someone asks you: "Do you know R?", you can now proudly and
confidently say: "Yes, some of it."
:::

-   The R universe is incredibly large and versatile. Whatever
    data-related task you can come up with - there is likely going to be
    an R-based solution to your issues.

-   New projects will likely require you to learn new things in R.

-   Luckily there are lots of resources to help you develop yourself as
    an R programmer.

### 

### Documentation

-   Cheat sheets:
-   `help()` and `?` in RStudio
-   Online package documentation (e.g. for
    [`ggplot2`](https://cran.r-project.org/web/packages/ggplot2/ggplot2.pdf))

### Stack Overflow (and other fora)

-   R users form a very active community.

-   Stack Overflow: Well-known forum and large archive for R questions
    (<https://stackoverflow.com/questions/tagged/r>)

-   Read replies for issues similar to yours or ask your own questions
    (remember to stick to the required format!)

-   Once you feel confident: Be an active part of the community.
    Contribute your own answers!

### AI 

-   AI tools are making coding easier and more convenient by the minute.
-   For instance, several AI tools can generate code based on prompts directly inside RStudio.
-   It is worth taking some time to consider the ethical implications of using AI in your work (e.g. https://www.unesco.org/en/artificial-intelligence/recommendation-ethics)

::: emphasis
Don't make the mistake to become a passive user. AI is only as good as the prompts you supply and it is not without its limitations. You'll need at least a basic understanding of what you are doingin R AND with the AI to accurately describe what you want and understand when the AI is making a mistake or misinterpreting your prompts.

Always double-check and test any results!
:::

###

#### Github Co-Pilot for R Studio

-   Github Co-Pilot is a new AI tool that can generate code based on prompts in form of comments directly inside RStudio.
-   It's free for UvA students.

::: question 
Step 1: Sign-up for Github Co-Pilot on Github
:::

::: question 
Step 2: Install the Github Co-Pilot extension in RStudio
:::

::: question 
Step 3: Start using Github Co-Pilot in RStudio
:::

###

Let's try it out.

::: question
Open an R Script and write the following comment: "# Add 3 to 5"
:::

::: question
Now a little more complicated: "# "
:::

```{r}
# A function that reports the result of t.test in APA format


```


[ADD exercise to write prompts for the AI and see how it does - do some
with the teacher and some on your own]

### Community

-   take and give back!

## Collaborate

various ways to collaborate - example GitHub

## Practice Open, Reproducible Science

-   Share data and code with DOI identifiers (e.g. github, OSF,
    Figshare)

-   Reproducible reports

-   Reproducible power points

## Fancy Stuff

### Shiny

The [`shiny`]() package is the RStudio contribution to interactive R
products. Actually, this tutorial is made with Shiny and it contains a
Shiny app, namely, the animated plots used in the *Evaluating a plot*
Section.

Here is the code for these plots. You cannot run the code here because
we cannot start a Shiny app from within a Shiny app (this tutorial). If
you create a new Shiny app in RStudio (*File \> New File \> Shiny Web
App*) and copy and paste the below code into the Shiny app file
(replacing all existing contents), you can run the app from RStudio (use
the *Run App* button).

::: question
Experiment with the user interface and the R code.
:::

```{r shiny1, eval=FALSE, echo=TRUE}
# load the shiny package
library(shiny)

# first part of the app: the User Interface (ui)
ui <- fluidPage(
    fluidRow( #the first (and only) row in the interface
        column(2, #the first (left) column, width 2 out of 12
               radioButtons("radio", label = h3("Select a plot"),
                            choices = list("Plot 1" = 1, #label and value if selected
                                           "Plot 2" = 2, 
                                           "Plot 3" = 3, 
                                           "Plot 4" = 4, 
                                           "Plot 5" = 5, 
                                           "Plot 6" = 6), 
                            selected = 1),
               #only a slider can be animated
               sliderInput("slider", label = "", min = 0, max = 20, value = 0,
                           step = 1, ticks = FALSE, 
                           animate = TRUE
               )
        ),
        column(10, #the second (right) column, width 10 out of 12
               plotOutput("regPlot") #containing the plot named regPlot
        )
    )  
)

# second part of the app: the R code
server <- function(input, output, session) {
  # create a plot named regPlot to be shown in the output
  output$regPlot <- renderPlot({
    # the basic (empty) plot  
    g <- ggplot(
      data = consumers,
      aes(x = ad_expo, y = brand_aw)
      ) +
      scale_x_continuous(name = "Exposure", breaks = 1:10, limits = c(1, 10)) +
      scale_y_continuous(name = "Brand awareness", breaks = 1:10, limits = c(1, 10)) +
      theme_bw( base_size = 14 )
    
    ## Create a plot version for each radio option
    if (input$radio == 1) { 
      # just add regression line and confidence interval to the empty plot
      g <- g + geom_smooth(method = "lm", 
                           formula = y ~ x, se = TRUE, 
                           color = "black")
    }
    else if (input$radio == 2) {
      # point size reflects brand_aw
      g <- g + 
        geom_point(aes(size = brand_aw), color = "grey") +
        geom_smooth(method = "lm", formula = y ~ x, se = FALSE)
    }
    else if (input$radio == 3) {
      # outlier not visible (and regression line without outlier)
      g <- g + 
        geom_point(size = 2, color = "grey") +
        geom_smooth(data = consumers[consumers$ad_expo > 1,], 
                    mapping = aes(x = ad_expo, y = brand_aw),
                    method = "lm", formula = y ~ x, se = FALSE) +
        scale_x_continuous(limits = c(4, 10))
    }
    else if (input$radio == 4) {
      # grey total regression line, red regression line without outlier
      g <- g + 
        geom_point(size = 2, 
          aes(color = ifelse(ad_expo > 1, "grey", "red")), show.legend = FALSE) +
        geom_smooth(method = "lm", formula = y ~ x, se = FALSE, 
                    color = "grey", size = 2) +
        geom_smooth(
          # trick: base R to create a subset of all consumers, omitting the outlier
          data = consumers[consumers$ad_expo > 1,], 
          mapping = aes(x = ad_expo, y = brand_aw),
          method = "lm", formula = y ~ x, se = FALSE, color = "red"
        ) +
        geom_text(x = 3.9, y = 4.3, label = "without outlier", 
                  size = 3, color = "red", alpha = 0.6) +
        scale_color_manual(values = c("red", "grey"))
    }
    else if (input$radio == 5) {
      # regression line per gender, with additional regression line 
      # for males without outlier
      g <- g + 
        geom_point(aes(color = gender), size = 2) +
        geom_smooth(
          method = "lm", formula = y ~ x, se = FALSE, aes(color = gender)
          ) +
        geom_smooth(
          data = consumers[consumers$ad_expo > 1 & consumers$gender == "male",], 
          mapping = aes(x = ad_expo, y = brand_aw),
          method = "lm", formula = y ~ x, se = FALSE, 
          color = "blue", linetype = "dashed"
        ) +
        geom_text(x = 4.1, y = 3.4, label = "without outlier", 
                  size = 3, color = "blue", alpha = 0.6)
    }
    else if (input$radio == 6) {
      # density conours and regression lines per gender
      g <- g + 
        geom_density2d(aes(color = gender),  adjust = 2) +
        geom_smooth(
          method = "lm", formula = y ~ x, se = FALSE, aes(color = gender)
          )
    }
    
    # Show plot, depending on slider value
    # This is the animation trick: slider values range from 0 to 20,
    # if it is 0, a text is shown, if it is 1 or 2 or at least 10, 
    # the plot is shown, a text is shown between 2 and 10, and
    # a text is added to the plot if the slider is 20.
    if (input$slider == 0) {
      #show press button text
      ggplot(data = consumers) + 
        geom_text(
          x = 0.5, y = 0.5, 
          label = "Press the little play button", size = 12
          ) 
    } 
    else if ((input$slider > 0 & input$slider < 4) | (input$slider > 10)) {
      # ask user to reset the slider 
      if (input$slider == 20) {
        if (input$radio == 3 ) {hor_pos = 7}
        else {hor_pos = 5.5}
        g <- g + geom_text(
          x = hor_pos, y = 8.5, 
          label = "Reset the slider to 0\nbefore you watch another plot.", 
          size = 8
          )
      }
      #show plot
      g
    } 
    else {
      # What did you see?
      ggplot(data = consumers) + 
        geom_text(
          x = 0.5, y = 0.5, 
          label = "What did you see?", size = 18
          )
    }
  })  
}

# the command to create and run the app
shinyApp(ui, server)
```

Check out the [Shiny demos
gallery](https://shiny.rstudio.com/gallery/#demos) for inspiration.
Start simple!

If you want to create a Shiny app, first create your plot with
`ggplot()`, then add it to the Shiny app.
