---
title: 'Session 14: Your Future with R'
output:
  learnr::tutorial:
    progressive: yes
    allow_skip: yes
runtime: shiny_prerendered
description: |
  Easy work flows and strategies to use R in collaborative, open science research.
---

```{r setup, include=FALSE}
# Ensure that libraries are loaded.
library(tidyverse)
library(learnr)
library(gradethis)
library(knitr)
library(kableExtra)
library(haven) #For importing SPSS data files.
library(car) #For ANOVA.  
library(texreg) #For pretty regression results.
library(effects) #For two-way interaction plots.
library(broom) #For cleaning up statistical results.

tutorial_options(exercise.timelimit = 20, exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r, context="data", include=FALSE}
# Ensure that the data is loaded for the remainder of this tutorial.
glbwarm <- UsingRTutorials::glbwarm
glbwarm_spss <- UsingRTutorials::glbwarm_spss
# The estimated regression model with rstanarm. 
model_1aBayes <- UsingRTutorials::model_1aBayes
```

<!-- Define programming tip style -->

```{=html}
<style>
.tip {
  background-color: #f5f5f5;
}
</style>
```
<!-- Define question style -->

```{=html}
<style>
.question {
  color: #5A9DDB;
}
</style>
```
<!-- Define emphasis style -->

```{=html}
<style>
.emphasis {
  color: #e8301b;
}
</style>
```

## Overview

## (Reference) Importing SPSS Data

### 

SPSS data files have a complicated setup with variable labels and value labels.

R data frames or tibbles do not have such labels.

In case you later have SPSS data that you want to analyze in R, here are **two options** for importing SPSS data.

### Option 1. Export from SPSS to .csv and import .csv in R.

-   Export data from SPSS to a CSV file with value labels for categorical variables:
    -   *File \> Export \> CSV Data* with *Save value labels where defined instead of data values*.
-   Use `read_csv()` (as you learned before) to import the CSV file.

::: question
Import the SPSS file `data/glbwarm.csv` and have a look at it.
:::

```{r importCSV, exercise = TRUE, exercise.eval = TRUE}
# The CSV is available in the data directory of this tutorial.
glbwarm <- read_csv("data/glbwarm.csv")
# Inspect the variables.
str(glbwarm)
```

### Option 2. Import SPSS .sav file directly with tidyverse package `haven`.

The tidyverse package `haven` contains function `read_sav()` (or `read_spss()`) for importing SPSS (and other software packages) data files.

::: question
Import the SPSS file `data/glbwarm.sav` and have a look at it.
:::

```{r importSPSS, exercise = TRUE}
# The SPSS system file is available in the data directory of this tutorial.
glbwarm_spss <- haven::read_sav("data/glbwarm.sav")
# Inspect the variables.
str(glbwarm_spss)
```

### 

Imported categorical variables such as *ideology*:

-   Are numerical codes with labels.
-   So, R treats these variables as numerical.
-   If you don't want that, change them into factors.

::: question
Use the `haven::as_factor()` function to add a variable named `sex_fct` to tibble `glbwarm_spss`.
:::

```{r sex_fct, exercise = TRUE}
# Add sex as a factor to the tibble.
glbwarm_spss <- glbwarm_spss |> 
  mutate(sex_fct = haven::as_factor(sex))
# Inspect the original and new variable.
glbwarm_spss |> count(sex, sex_fct)
```

### 

Passing the entire tibble `glbwarm_spss` to `haven::as_factor()`will change all labelled variables into factors.

I am not sure this is what you want here. Perhaps you would like to use the `ideology` variable as a seven-point scale.

### Shiny

The [`shiny`]() package is the RStudio contribution to interactive R
products. Actually, this tutorial is made with Shiny and it contains a
Shiny app, namely, the animated plots used in the *Evaluating a plot*
Section.

Here is the code for these plots. You cannot run the code here because
we cannot start a Shiny app from within a Shiny app (this tutorial). If
you create a new Shiny app in RStudio (*File \> New File \> Shiny Web
App*) and copy and paste the below code into the Shiny app file
(replacing all existing contents), you can run the app from RStudio (use
the *Run App* button).

::: question
Experiment with the user interface and the R code.
:::

```{r shiny1, eval=FALSE, echo=TRUE}
# load the shiny package
library(shiny)

# first part of the app: the User Interface (ui)
ui <- fluidPage(
    fluidRow( #the first (and only) row in the interface
        column(2, #the first (left) column, width 2 out of 12
               radioButtons("radio", label = h3("Select a plot"),
                            choices = list("Plot 1" = 1, #label and value if selected
                                           "Plot 2" = 2, 
                                           "Plot 3" = 3, 
                                           "Plot 4" = 4, 
                                           "Plot 5" = 5, 
                                           "Plot 6" = 6), 
                            selected = 1),
               #only a slider can be animated
               sliderInput("slider", label = "", min = 0, max = 20, value = 0,
                           step = 1, ticks = FALSE, 
                           animate = TRUE
               )
        ),
        column(10, #the second (right) column, width 10 out of 12
               plotOutput("regPlot") #containing the plot named regPlot
        )
    )  
)

# second part of the app: the R code
server <- function(input, output, session) {
  # create a plot named regPlot to be shown in the output
  output$regPlot <- renderPlot({
    # the basic (empty) plot  
    g <- ggplot(
      data = consumers,
      aes(x = ad_expo, y = brand_aw)
      ) +
      scale_x_continuous(name = "Exposure", breaks = 1:10, limits = c(1, 10)) +
      scale_y_continuous(name = "Brand awareness", breaks = 1:10, limits = c(1, 10)) +
      theme_bw( base_size = 14 )
    
    ## Create a plot version for each radio option
    if (input$radio == 1) { 
      # just add regression line and confidence interval to the empty plot
      g <- g + geom_smooth(method = "lm", 
                           formula = y ~ x, se = TRUE, 
                           color = "black")
    }
    else if (input$radio == 2) {
      # point size reflects brand_aw
      g <- g + 
        geom_point(aes(size = brand_aw), color = "grey") +
        geom_smooth(method = "lm", formula = y ~ x, se = FALSE)
    }
    else if (input$radio == 3) {
      # outlier not visible (and regression line without outlier)
      g <- g + 
        geom_point(size = 2, color = "grey") +
        geom_smooth(data = consumers[consumers$ad_expo > 1,], 
                    mapping = aes(x = ad_expo, y = brand_aw),
                    method = "lm", formula = y ~ x, se = FALSE) +
        scale_x_continuous(limits = c(4, 10))
    }
    else if (input$radio == 4) {
      # grey total regression line, red regression line without outlier
      g <- g + 
        geom_point(size = 2, 
          aes(color = ifelse(ad_expo > 1, "grey", "red")), show.legend = FALSE) +
        geom_smooth(method = "lm", formula = y ~ x, se = FALSE, 
                    color = "grey", size = 2) +
        geom_smooth(
          # trick: base R to create a subset of all consumers, omitting the outlier
          data = consumers[consumers$ad_expo > 1,], 
          mapping = aes(x = ad_expo, y = brand_aw),
          method = "lm", formula = y ~ x, se = FALSE, color = "red"
        ) +
        geom_text(x = 3.9, y = 4.3, label = "without outlier", 
                  size = 3, color = "red", alpha = 0.6) +
        scale_color_manual(values = c("red", "grey"))
    }
    else if (input$radio == 5) {
      # regression line per gender, with additional regression line 
      # for males without outlier
      g <- g + 
        geom_point(aes(color = gender), size = 2) +
        geom_smooth(
          method = "lm", formula = y ~ x, se = FALSE, aes(color = gender)
          ) +
        geom_smooth(
          data = consumers[consumers$ad_expo > 1 & consumers$gender == "male",], 
          mapping = aes(x = ad_expo, y = brand_aw),
          method = "lm", formula = y ~ x, se = FALSE, 
          color = "blue", linetype = "dashed"
        ) +
        geom_text(x = 4.1, y = 3.4, label = "without outlier", 
                  size = 3, color = "blue", alpha = 0.6)
    }
    else if (input$radio == 6) {
      # density conours and regression lines per gender
      g <- g + 
        geom_density2d(aes(color = gender),  adjust = 2) +
        geom_smooth(
          method = "lm", formula = y ~ x, se = FALSE, aes(color = gender)
          )
    }
    
    # Show plot, depending on slider value
    # This is the animation trick: slider values range from 0 to 20,
    # if it is 0, a text is shown, if it is 1 or 2 or at least 10, 
    # the plot is shown, a text is shown between 2 and 10, and
    # a text is added to the plot if the slider is 20.
    if (input$slider == 0) {
      #show press button text
      ggplot(data = consumers) + 
        geom_text(
          x = 0.5, y = 0.5, 
          label = "Press the little play button", size = 12
          ) 
    } 
    else if ((input$slider > 0 & input$slider < 4) | (input$slider > 10)) {
      # ask user to reset the slider 
      if (input$slider == 20) {
        if (input$radio == 3 ) {hor_pos = 7}
        else {hor_pos = 5.5}
        g <- g + geom_text(
          x = hor_pos, y = 8.5, 
          label = "Reset the slider to 0\nbefore you watch another plot.", 
          size = 8
          )
      }
      #show plot
      g
    } 
    else {
      # What did you see?
      ggplot(data = consumers) + 
        geom_text(
          x = 0.5, y = 0.5, 
          label = "What did you see?", size = 18
          )
    }
  })  
}

# the command to create and run the app
shinyApp(ui, server)
```

Check out the [Shiny demos
gallery](https://shiny.rstudio.com/gallery/#demos) for inspiration.
Start simple!

If you want to create a Shiny app, first create your plot with
`ggplot()`, then add it to the Shiny app.

