---
title: 'Session 6: Modelling'
output:
  learnr::tutorial:
    progressive: yes
    allow_skip: yes
runtime: shiny_prerendered
description: |
  We don't follow the approach to data analysis presented in the book. Instead, we apply our skills in statistical analysis within R.
---

```{r setup, include=FALSE}
# Ensure that libraries are loaded.
library(tidyverse)
library(learnr)
library(gradethis)
library(knitr)
library(kableExtra)
# New packages (must be installed before taking this tutorial)
# library(haven) #For importing SPSS data files, part of tidyverse.
library(car) #For ANOVA.  
# If you want to create a regression table or moderation plot, install and load these packages.
library(texreg) #For pretty regression results.
library(effects) #For two-way interaction plots.

tutorial_options(exercise.timelimit = 60, exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r, context="data", include=FALSE}
# Ensure that the data is loaded for the remainder of this tutorial.
glbwarm <- UsingRTutorials::glbwarm
glbwarm_spss <- UsingRTutorials::glbwarm_spss
```

<!-- Define programming tip style -->
<style>
.tip {
  background-color: #f5f5f5;
}
</style>

<!-- Define question style -->
<style>
.question {
  color: #5A9DDB;
}
</style>

## Import SPSS Data

### 

SPSS data files have a complicated setup with variable labels and value labels. R data frames do not have such labels. 

Let us have a look at two options for importing SPSS data into R.

Option 1. Export from SPSS to .csv and import .csv in R.

- Export data from SPSS to a CSV file with value labels for categorical variables: _File > Export > CSV Data_ with _Save value labels where defined instead of data values_.
- Use `read_csv()` (as you learned before) to import the CSV file. 

The resulting R data set is available in this tutorial as `glbwarm`.

<div class="question">
Have a look at the imported CSV file. Any variables that had better become factors (and why)?
</div>

```{r importCSV, exercise = TRUE}

```

```{r importCSV-hint-1}
str(glbwarm)
```

###

Option 2. Import SPSS .sav file directly with tidyverse package `haven`.

The resulting R data set is available in this tutorial as `glbwarm_spss`.

<div class="question">
Have a look at the imported SPSS file. Use R code to retrieve the label of variable `govact`.
</div>

```{r importSPSS, exercise = TRUE}

```

<!-- To hide the solution, use a textual hint. -->
<div id="importSPSS-hint">
__Hint:__ You learned how to retrieve a label from a list in Session 5.
</div>

```{r importSPSS-solution}
attributes(glbwarm_spss$govact)$label
```

```{r importSPSS-check}
gradethis::grade_code(
  correct = "", 
  incorrect = ""
  )
```

###

The numerical codes for categorical variables such as ideology in the SPSS file have been preserved. 

R treats these variables as numerical. Maybe this is what you want in some instances. If you do not want this, you should change variables into factors. 

<div class="question">
Use the `haven::as_factor()` function to add a variable named `sex_fct` to tibble `glbwarm_spss`.
</div>

```{r sex_fct, exercise = TRUE}

```

<!-- To hide the solution, use a textual hint. -->
<div id="sex_fct-hint">
__Hint:__ Both the `forcats` and `haven` package contain a function `as_factor()`, so it is recommended to specify the package when using this function.
</div>

```{r sex_fct-solution}
glbwarm_spss <- glbwarm_spss %>% mutate(sex_fct = haven::as_factor(glbwarm$sex))
```

```{r sex_fct-check}
gradethis::grade_code(
  correct = "message", 
  incorrect = "message"
  )
```

###

Note that passing the entire tibble `glbwarm_spss` to `haven::as_factor()` will change all labelled variables into factors. I am not sure this is what you want here. Perhaps you would like to use the `ideology` variable as a seven-point scale.

## Basic Statistical Models

Let us practice with some of the most common statistical analyses in R. 

If you need help, consult Sections 3 and 4 in [_Help, My Collaborator Uses R! An Introduction to Reproducible Statistical Analyses in R_](https://wdenooy.github.io/Switch2R/index.html) and R help on the functions that we use.

### t Test: `t.test()`

You already know how to execute an independent-samples t test (Session 5). There are different versions of the same function for different t tests.

Vignette: 

```{r t.test.vignette, eval=FALSE, echo=TRUE}
t.test(x, ...)

### Default S3 method:
t.test(x, y = NULL,
       alternative = c("two.sided", "less", "greater"),
       mu = 0, paired = FALSE, 
       var.equal = FALSE,
       conf.level = 0.95, ...)

### S3 method for class 'formula'
t.test(formula, data, subset, na.action, ...)
```

- the function with just `x` is for a one sample t test: specify the hypothesized population mean with argument `mu =`.  
- the function with `x, y` is for paired samples t tests.  
- the function with a formula is for two samples t tests; `y` must be a variable with two categories.  

### F test on Two Variances: `var.test()`  

In contrast to SPSS, R only gives you what you ask for. If you ask for a t test, you get a t test but not checks on assumptions. You have to apply those checks yourself.

You may remember that the version of the independent samples t test that we must use depends on whether the population variances are equal for the two groups. 

<div class="question">
Use the function `var.test` to test if `govact` variances are equal for females and males in the population. Store the results as a data object named `vartest`.
</div>

```{r vartest1, exercise = TRUE}

```

<!-- To hide the solution, use a textual hint. -->
<div id="vartest1-hint">
__Hint:__ Have a look at the help for function `var.test`.
</div>

```{r vartest1-solution}
vartest <- var.test(govact ~ sex, data = glbwarm)
```

```{r vartest1-check}
gradethis::grade_code(
  correct = "", 
  incorrect = ""
  )
```

<div class="tip" >
__Remember__

- R formula: dependent variable/outcome ~ independent variable/predictor (and more).
</div>

###

<div class="question">
Pull the p value from data object `vartest` that you have just created. Is the test on equal population variances statistically significant?
</div>

```{r vartestp, exercise = TRUE, exercise.setup = "vartest1-solution"}

```

<!-- To hide the solution, use a textual hint. -->
<div id="vartestp-hint">
__Hint:__ Review Session 5 if you do not know how to do this.
</div>

```{r vartestp-solution}
vartest$p.value
```

```{r vartestp-check}
gradethis::grade_code(
  correct = "`e-08` means `* 10^-8` (scientific notation). Note that the results are stored as class htest, just like the results from `t.test()`.", 
  incorrect = "Perhaps you used double square brackets instead of the dollar sign to pull out the p value. That's OK."
  )
```

###

In the R code for an independent samples t test, the `var-equal` argument must be set to `TRUE` or `FALSE`. We can 

<div class="question">
Integrate the F test on equal population variances in the t test, such that the t test automatically uses the correct version: with or without equal population variances assumed. Send the results to the screen (do not save it as a data object).
</div>

```{r ttest2, exercise = TRUE}

```

<!-- To hide the solution, use a textual hint. -->
<div id="ttest2-hint">
__Hint:__ Arguments that are at their default values are not included in the answer that is used for checking.
</div>

```{r ttest2-solution}
t.test(govact ~ sex, data = glbwarm, var.equal = var.test(govact ~ sex, data = glbwarm)$p.value > 0.05)
```

```{r ttest2-check}
gradethis::grade_code(
  correct = "", 
  incorrect = ""
  )
```

### Linear Regression: `lm()`  

Vignette:
```{r lm_vignette, eval=FALSE, echo=TRUE}
lm(formula, data, subset, weights, na.action,
   method = "qr", model = TRUE, x = FALSE, y = FALSE, qr = TRUE,
   singular.ok = TRUE, contrasts = NULL, offset, ...)
```

See book p. 358-371 {Section 23.4} for using regression formulas, including interactions and transformations within a formula.

<div class="question">
Use `lm()` and tibble `glbwarm` to predict support for governmental action (`govact`) from age, negative emotions and party identification. Store the results in data object `model_1`.
</div>

```{r lm1, exercise = TRUE}

```

<!-- To hide the solution, use a textual hint. -->
<div id="lm1-hint">
__Hint:__ Use a formula.
</div>

```{r lm1-solution}
model_1 <- lm(govact ~ age + negemot + partyid, data = glbwarm)
```

```{r lm1-check}
gradethis::grade_code(
  correct = "", 
  incorrect = ""
  )
```

###

Data objects for results of statistical analyses always have a summary() and print() function for quick inspection. Not for presentation of results!


<div class="question">
Inspect the regression results with `summary()` and `print()`. What happened to the character variable?
</div>

```{r lm2, exercise = TRUE}

```

### Linear Regression: Two-Way Interaction  

Just like `lm()` takes care of creating dummies/indicator variables for a categorical predictor, it may also create interaction variables for you.

<div class="question">
Add an interaction effect between negative emotions (numeric) and age (numeric, in decades) to the regression model and save the results as data object `model_1a`.
</div>

```{r lm3, exercise = TRUE}

```

<!-- To hide the solution, use a textual hint. -->
<div id="lm3-hint">
__Hint:__ An interaction in a regression formula yields the partial effects of the individual variables and their interaction effect(s).
</div>

```{r lm3-solution, exercise.setup = "lm1-solution"}
model_1a <- lm(govact ~ age*negemot + partyid, data = glbwarm)
```

```{r lm3-check}
gradethis::grade_code(
  correct = "", 
  incorrect = ""
  )
```

###

<div class="question">
Replace the interaction effect in the previous model by an interaction effect between negative emotions (numeric) and party identification (categorical). Send the results to the screen. Can you make sense of the coefficients?
</div>

```{r lm4, exercise = TRUE}

```

<!-- To hide the solution, use a textual hint. -->
<div id="lm4-hint">
__Hint:__ An interaction in a regression formula yields the partial effects of the individual variables and their interaction effect(s).
</div>

```{r lm4-solution}
lm(govact ~ age + negemot*partyid, data = glbwarm)
```

```{r lm4-check}
gradethis::grade_code(
  correct = "", 
  incorrect = ""
  )
```

### Analysis of Variance: `lm()` and `car::Anova()`

In R, analysis of variance consists of two steps.

Step 1: ANOVA is linear regression with special contrasts (`contr.sum`).

Contrast `contr.sum` gives deviations from the mean. In analysis of variance, (main) effects are deviations from the (grand) mean.

<div class="question">
Estimate a regression model with support for governmental action (`govact`) predicted from respondent's sex and party identification, and the interaction between the two predictors. Use `contr.sum` contrasts and save the results as data object `model_2`.
</div>

```{r anova1, exercise = TRUE}

```

<!-- To hide the solution, use a textual hint. -->
<div id="anova1-hint">
__Hint:__ The `contrasts` argument requires a list of variable name and contrast type pairs.
</div>

```{r anova1-solution}
model_2 <- lm(govact ~ sex * partyid, data = glbwarm, contrasts=list(sex=contr.sum, partyid=contr.sum))
```

```{r anova1-check}
gradethis::grade_code(
  correct = "Have a look at the resuts.", 
  incorrect = ""
  )
```

###

Step 2: Calculate the sums of squares partition.

Functions:

- `stats::anova()` for balanced designs.
- `car::Anova()` for balanced and unbalanced designs.  

<div class="question">
Use the `Anova()` function to show the sums of squares partition with associated F tests on the screen.
</div>

```{r anova2, exercise = TRUE, exercise.setup = "anova1-solution"}

```

<!-- To hide the solution, use a textual hint. -->
<div id="anova2-hint">
__Hint:__ You may have to load the `car` package.
</div>

```{r anova2-solution}
Anova(model_2)
```

```{r anova2-check}
gradethis::grade_code(
  correct = "", 
  incorrect = "Perhaps you used the package name in the command, which is fine."
  )
```

###

The anova functions return a data frame, which you can use as any data frame. For example, knit it to a pretty (HTML or PDF) table with `knitr::kable()`. We will do that later on in this tutorial.

### Missing Values  

Statistical functions handle missing values correctly: 

- Cases with a missing value on a variable included in the analysis are dropped (listwise deletion). 
- Don't drop cases with missing values yourself with `na.omit()` because you may also drop cases with missing values only on variables that are not included in the analysis.  

## Print-Quality Results Tables

### Off-The-Shelf Tables

There are several packages that help you to tabulate statistical results. The table below lists some of them with their characteristic features.

```{r packageTables, echo=FALSE}
# Create a data frame for the contents of the table.
dt <- data.frame(
  Package = c("base, stats", "papaja", "stargazer", "texreg"),
  Models = c("all", "t test, regression, anova", "regression", "regression"),
  Format = c("plain text", "PDF, Word (HTML)", "PDF, HTML, plain", "PDF, HTML, plain"),
  Style = c("-", "APA6", "div., not APA6", "generic"),
  Comparison = c("-", "stacked", "side-by-side", "side-by-side"),
  Peculiarities = c("summary() and print(), only for quick inspection", "2 steps: apa_print() and apa_table()", "", "texreg(), hmtlreg(), screenreg()"),
  stringsAsFactors = FALSE
  )
names(dt)[5] <- paste0(names(dt)[5], footnote_marker_symbol(1))
dt %>%
  knitr::kable(align = "llllll", escape = F) %>% #show with kable() from the knitr package
  kable_styling(full_width = T) %>%
  row_spec(0, font_size = 28) %>%
  footnote(symbol = "Results of two of more models in one table.")
```

### Installing `papaja`

Package `papaja` is not in the CRAN repository (or any of its mirrors).

<div class="question">
Install `papaja` from GitHub (you must have internet connection) in RStudio.
</div>

```{r papaja, echo = TRUE, eval=FALSE }
# Execute this line of code in the RStudio console. 
devtools::install_github("crsh/papaja")
```

See https://crsh.github.io/papaja_man/ for a papaja manual.

We will see more of `papaja` in Session 7. Ensure that it is installed now in RStudio.

### Print-Quality Table With `texreg`

```{r model2_texreg, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(texreg)
model_1 <- lm(govact ~ age + negemot + partyid, data = glbwarm)
model_1a <- lm(govact ~ age*negemot + partyid, data = glbwarm)
# Table for HTML output.
texreg::htmlreg(list(model_1,model_1a), #the regression model(s) shown
        single.row = T, #coefficient and standard error on the same row
        star.symbol = "\\*", 
        doctype = F, #better for Markdown document
        html.tag = F, #better for Markdown document 
        head.tag = F, #better for Markdown document
        body.tag = F, #better for Markdown document 
        caption = "", #no caption to save space on the slide
        custom.coef.names = c(NA, "Age", "Negative emotions", "Independent", "Republican", "Age*Neg. emotions"),
        vertical.align.px = 6)
# For PDF output, use the texreg() function, with slightly different arguments (options).
# Use Help to see more arguments.
```

<div class="question">
The above table is generated from the code below. What happens if you run the code?
</div>

```{r texreg, exercise = TRUE, exercise.setup = "lm3-solution"}
# Attach the texreg package.
library(texreg)
# Table for HTML output.
texreg::htmlreg(list(model_1,model_1a), #the regression model(s) shown
        single.row = T, #coefficient and standard error on the same row
        star.symbol = "\\*", 
        doctype = F, #better for Markdown document
        html.tag = F, #better for Markdown document 
        head.tag = F, #better for Markdown document
        body.tag = F, #better for Markdown document 
        caption = "", #no caption to save space on the slide
        custom.coef.names = c(NA, "Age", "Negative emotions", "Independent", "Republican", "Age*Neg. emotions"),
        vertical.align.px = 6)
# For PDF output, use the texreg() function, with slightly different arguments (options).
# Use Help to see more arguments.
```

###

`htmlreg()` produces HTML code. This code should not be treated as ordinary text when the RMarkdown document is knitted. Instead, it must be used and formatted as HTML code. 

The `results='asis'` code chunk option is needed to get html as html instead of html as text output from R. 

This is how the code looks in the RMarkdown document.

````markdown
`r ''````{r eval=TRUE, echo=FALSE, results='asis'}
# Attach the texreg package.
library(texreg)
# Table for HTML output.
texreg::htmlreg(list(model_1,model_1a), #the regression model(s) shown
        single.row = T, #coefficient and standard error on the same row
        star.symbol = "\\*", 
        doctype = F, #better for Markdown document
        html.tag = F, #better for Markdown document 
        head.tag = F, #better for Markdown document
        body.tag = F, #better for Markdown document 
        caption = "", #no caption to save space on the slide
        custom.coef.names = c(NA, "Age", "Negative emotions", "Independent", "Republican", "Age*Neg. emotions"),
        vertical.align.px = 6)
# For PDF output, use the texreg() function, with slightly different arguments (options).
# Use Help to see more arguments.
```
````

###

Note that the `texreg` package has different functions with partly different options for different output formats.

`htmlreg()` is one example of a function that operates on lm() objects. Also see `confint()`, `coef()`, `resid()`.

<div class="question">
Find out what these functions do. Apply them to `model_1` and check out the options of these functions.
</div>

```{r lmfunctions, exercise = TRUE}

```

<!-- To hide the solution, use a textual hint. -->
<div id="lmfunctions-hint">
__Hint:__ Read the help info to these functions.
</div>

### Custom Tables with `broom` and `knitr`

If you want to have full control of your table, create it yourself with packages `broom` and `knitr`. Note that `broom` is part of tidyverse, so you do not have to load it separately.

Step 1: Function `broom::tidy()` extracts relevant statistics from a statistical results object into a tibble.

<div class="question">
Use `broom` and data objext `model_1a` to create a tibble named `results_1` containing the regression coefficients with their standard errors, t values, p values, and 95% confidence intervals.
</div>

```{r broom, exercise = TRUE}

```

<!-- To hide the solution, use a textual hint. -->
<div id="broom-hint">
__Hint:__ Check out help on `broom::tide.lm()`. You are tidying the results of a linear model (`lm()`).
</div>

```{r broom-solution}
results_1 <- broom::tidy(model_1a, conf.int = TRUE, conf.level = 0.95)
```

```{r broom-check}
gradethis::grade_code(
  correct = "Send results_1 to the screen to see the results.", 
  incorrect = "Perhaps you did not specify the `broom` package with the `tidy()` function (`broom::tidy()`), which is OK."
  )
```

###

`broom` produces a tibble (data frame), so you can manipulate it like any other.

Step 2: Select and adjust values to suit your needs.

<div class="question">
Change tibble `results_1` into tibble `results_2`, which contains the effect (variable) names, the unstandardized regression coefficients, p values, and the confidence intervals (named `CI`) in APA format (e.g., `[-0.56, 0.11]`).
</div>

```{r broom2, exercise = TRUE, exercise.setup = "broom-solution"}

```

<!-- To hide the solution, use a textual hint. -->
<div id="broom2-hint">
__Hint:__ Use your data wrangling skills. Use `paste0()` to get the confidence interval in the right format.
</div>

```{r broom2-solution, exercise.setup = "broom-solution"}
results_2 <- results_1 %>% mutate(CI = paste0( "[", format( round(conf.low, digits = 2), nsmall = 2 ), ", ", format( round(conf.high, digits = 2), nsmall = 2 ), "]" )) %>% select(term, estimate, p.value, CI)
```

```{r broom2-check}
gradethis::grade_code(
  correct = "", 
  incorrect = "Send the resulting tibble to the screen to check your results. Maybe it is OK, but you just used a different order for the commands."
  )
```

###

If you want to use stars to mark the significance level of regression coefficients, you can add a new character variable showing the number of stars.

<div class="question">
Create a new character variable named `sig` that contains the number of stars indicating the significance level. Save the effect ame, regression coefficient, number of stars, and the confidence interval (in this order) in the data object `results_3`.
</div>

```{r broom3, exercise = TRUE, exercise.setup = "broom2-solution"}

```

<!-- To hide the solution, use a textual hint. -->
<div id="broom3-hint">
__Hint:__ Use `case_when()` and `p.value` to create the new `sig` variable.
</div>

```{r broom3-solution }
results_3 <- results_2 %>% mutate( sig = case_when( p.value < .001 ~ "***", p.value < .01 ~ "**", p.value < .05 ~ "*", TRUE ~ "" )) %>% select(term, estimate, sig, CI)
```

```{r broom3-check}
gradethis::grade_code(
  correct = "", 
  incorrect = ""
  )
```

###

Step 3: Create a table with `knitr::kable()` and `kableExtra`

With `knitr` and `kableExtra`, we can create a table including footnotes.

<div class="question">
Play around with the `kable` and `kableExtra` options to see what they do.
</div>

```{r kablehidden, eval=TRUE, echo=FALSE}
# Ensure that results_3 exists.
results_3 <- broom::tidy(model_1a, conf.int = TRUE, conf.level = 0.95) %>%
  mutate(CI = paste0( "[", format( round(conf.low, digits = 2), nsmall = 2 ), ", ", format( round(conf.high, digits = 2), nsmall = 2 ), "]" )) %>%
  select(term, estimate, p.value, CI) %>%
  mutate(
    sig = case_when(
      p.value < .001 ~ "***",
      p.value < .01 ~ "**",
      p.value < .05 ~ "*",
      TRUE ~ ""
    )
  )  %>%
  select(term, estimate, sig, CI)
```


```{r kable, exercise = TRUE, exercise.eval = TRUE, exercise.lines = 17}
results_3 %>% kable(digits = c(0, 2, 0, 0),
  col.names = c("Parameter", "B", "", "95% CI"),
  align = "lrlc",
  caption = "Table 1. Predicting opinions about global warming.",
  booktabs = TRUE, #nicer layout in PDF
  escape = FALSE #pay attention to special characters
  ) %>%
  kable_styling(full_width = FALSE) %>%
  row_spec(0, font_size = 16) %>%
  column_spec(1, width = "5cm") %>%
  column_spec(2, width = "3cm") %>%
  column_spec(3, width = "0.5cm") %>%
  column_spec(4, width = "5cm") %>%
  footnote(
    general_title = "",
    general = "   * p < .05. ** p < .01. *** p < .001."
    )
```

###

Some final points about tabulating results:

- Special characters such as stars (`*`) and percentage signes (`%`) can be troublesome in tables. Sometimes, you have to escape them with one or more backslashes (`\\`).

- PDF output has more formatting options than HTML (or Word).

- `kable()` does not knit nicely to Word. Knit to HTML and import HTML in Word.

## Results Plots

A data object with statistical results usually has a `plot()` function. These plots are for quick inspection rather than final presentation. They can be very useful for checking assumtpions.

<div class="question">
Apply the `plot()` function to the result of linear regression (`model_1a`). 

- Which plots do you get? 
- Are these all plots that you can get with this function?
</div>

```{r lmplots, exercise = TRUE}

```

<!-- To hide the solution, use a textual hint. -->
<div id="lmplots-hint">
__Hint:__ See `plot.lm()` for help.
</div>

### Off-The-Shelf Plots

There are many packages offering ready-to-use plots, for example:

- `papaja`: plots for analysis of variance.
- `coefplot`: plots regression coefficients for one or more models (`ggplot2` plots)
- `visreg`: plots regression lines (`ggplot2` plots).
- `effects`: plots regression lines (not `ggplot2` plots).

Note: `ggplot2` plots can be further customized: save the plot (e.g., `p`) and then add layers, themes, ... (e.g., `p + theme_bw()`). 

### Interaction Plot with the `effects` Package

You can graph interaction effects with the `effects` package in two steps.

```{r effects, exercise=TRUE, exercise.setup = "lm3-solution", exercise.eval = TRUE }
# Load effects package.
library(effects)
# Step 1: Create a data object containing all effects.
eff.model2 <- effects::allEffects(model_1a)
# Step 2: Plot interaction effects.
effects::plot(eff.model2, 'age:negemot', x.var = 'age')
```

Note the `rug` on the horizontal axis, showing the age score of all cases within a `negemot` group. Good coverage of the variable on the x axis is required for reliable conditional regression lines.  

### Custom Plots with `ggplot()`

It is not so difficult to create this plot with `ggplot()`. If you do that, you can  control the classes on negative emotions that you want to use. For example, why does the plot from the `effects` package skip negative emotions around three?
  
<div class="question" >
Create a ggplot from `glbwarm` like the with a facets for negative emotions between 1 and 1.5 (labeled `1`), between 1.5 and 2.5 (labeled `2`), between 2.5 and 3.5 (labeled `3`), between 3.5 and 4.5 (labeled `4`), between 4.5 and 5.5 (labeled `5`), between 5.5 and 6 (labeled `6`). Name the new variable `negemot_bin`.
</div>

```{r ggplot, exercise = TRUE, exercise.setup = "lm3-solution"}

```

```{r ggplot-hint-1}
# Create the binned negative emotions variable.
glbwarm %>%
  mutate(negemot_bin = 
  case_when(
    negemot < 1.5 ~ 1,
    negemot < 2.5 ~ 2,
    negemot < 3.5 ~ 3,
    negemot < 4.5 ~ 4,
    negemot < 5.5 ~ 5,
    negemot >= 5.5 ~ 6
    )
  )
```

```{r ggplot-hint-2}
# Use geom_rug() to represent all observations on the horizontal axis.
```

```{r ggplot-hint-3}
# Use facet_wrap() on the binned negative emotions variable.
```

```{r ggplotresult, eval = TRUE, echo = FALSE}
a <- glbwarm %>%
  mutate(negemot_bin = 
  case_when(
    negemot < 1.5 ~ 1,
    negemot < 2.5 ~ 2,
    negemot < 3.5 ~ 3,
    negemot < 4.5 ~ 4,
    negemot < 5.5 ~ 5,
    negemot >= 5.5 ~ 6
    )
  ) %>%
  ggplot(aes(x = age)) +
    geom_smooth(aes(y = govact), method = lm) +
    geom_rug() +
    facet_wrap(vars(negemot_bin))
```

Do you notice differences between your plot and the plot created with the `effects` package. Which plot do you trust more?

### 

It is not that difficult to create a means plot showing the results of analysis of variance.

```{r anovaplothidden, eval=TRUE, echo=FALSE}
glbwarm %>% group_by(partyid, sex) %>% 
  summarise(avg_govact = mean(govact)) %>% 
  ggplot(aes(x = partyid, y = avg_govact, 
             color = sex)) + 
    geom_line(aes(group = sex)) + 
    geom_point() +
    labs(x = "Party identification",
    y = "Gov.intervention") +
    scale_y_continuous(
      limits = c(min(glbwarm$govact), max(glbwarm$govact)),
      breaks = 1:7
    ) +
    theme_bw() +
    theme(legend.position = c(0.8, 0.8),
      legend.background = element_blank())
```

<div class="question">
Use `gpglot()` to create the above means plot.
</div>

```{r anovaplot, exercise = TRUE}

```

```{r anovaplot-hint-1}
# First calculate the group means that must be shown.
```

```{r anovaplot-hint-2}
# Use geom_point() to show the dots.
```

```{r anovaplot-hint-3}
# Use geom_line() to show the lines with the group argument.
```

```{r anovaplot-hint-4}
# Use theme_bw() for the general appearance of the plot.
# More on this in Session 7.
```

```{r anovaplot-hint-5}
# Use legend.position and legend.background within theme() for the fine details of the legend.
```

## Data Project

No fun stuff in this tutorial. Work on your Data Project.

Statistical analyses are not necessary for the Data Project. The Data Project focuses on visualizations.

You can, however, use statistical analysis to detect patterns in your data that you then try to visualize. If you do that, do not use off-the-shelf plots. Show that you can create a plot that hopefully is more attractive and more informative than off-the-shelf statistical plots.
