---
title: "Session 3: Data Tidying (1)"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
    Check your understanding of the tidyverse functions for managing complex datasets, focusing on tidy data principles 1 and 2, 
    learn how to wrangle in steps and check each step,
    and continue working on your Data Project.
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
# Ensure that libraries are loaded.
library(tidyverse)
library(learnr)
library(gradethis)
library(knitr)
library(kableExtra)

tutorial_options(exercise.timelimit = 60, exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r, context="data", include=FALSE}
# Ensure that the data is loaded for the remainder of this tutorial.
GlasgowFriends <- UsingRTutorials::GlasgowFriends # readr::read_csv("data/GlasgowFriends.csv")
# Create the Student table.
# It is not used in grade_result(), so it may already be around.
Student <- GlasgowFriends |> select(student, neighbourhood, schooldist, hoodname, age, sex, smoking_at_home, smoking_parents, smoking_siblings) |> distinct()
```

<!-- Define programming tip style -->

```{=html}
<style>
.tip {
  background-color: #f5f5f5;
}
</style>
```
<!-- Define question style -->

```{=html}
<style>
.question {
  color: #5A9DDB;
}
</style>
```
<!-- Define emphasis style -->

```{=html}
<style>
.emphasis {
  color: #e8301b;
}
</style>
```
## Overview

-   AI
-   Four principles of tidy data
-   Principle 1: Each type of case must have its own tibble
-   Principle 2: Each observation must have it's own row

## AI

-   AI can be a great coding help, ONCE you have a basic understanding
    of R.
-   When you try to use AI before you understanding the basics:
    -   You will likely write low quality prompts.
    -   You are unlikely to understand and critically evaluate the
        output.
-   Our recommendation: Learn on your own first. Use AI later to keep
    developing (e.g. using GitHub Co-Pilot for R).
-   In this class, we will not use AI to generate code. The exam will be
    closed book and without internet connection.

## Tidy Data: Eve`R`ything In One Place

-   Rules for optimal data structure designed for efficient and
    consistent data storage.
-   Key goals are to facilitate easy storage and (re-)use of complex
    datasets by
-   Storing data in a way that conserves space,
-   Storing data in a way that is easily understandable by others,
-   Preparing the dataset for common operations (e.g. merging different
    types of data) given the characteristics of the R programming
    language.

## Four Principles of Tidy Data

1.  Each type of case must have its own tibble (table).
    -   Is the same information repeated in different rows?\
    -   Split the tibble; transfer the info on each type of case to a
        separate tibble.\
    -   The book *R for Data Science (2nd Edition)* does not discuss
        this much. See Chapter 19 on Joins.
2.  Each observation must have its own row.
    -   Do different variables address the same information?\
    -   Stack repeated information.\
    -   Don't allow values as columns.
3.  Each variable must have its own column.
    -   Is there a variable with values that should be variables by
        themselves?
    -   Don't stack different types of information in the same column.
    -   Spread information over different columns.
4.  Each value must have its own cell.
    -   Does a variable contain more than one piece of information?
    -   Don't cram more than one piece of information in a cell.\
    -   Split the variable.

Today, we will focus on principle 1 and 2. In the next session, we'll
focus on principles 3 and 4.

## 1. Each type of case must have its own tibble

In this tutorial, we tidy up the very untidy tibble `GlasgowFriends`.

Data arising from practical sources usually contain information about
different types of cases, for example:

-   a person;
-   an organization;
-   a moment in time;
-   or a combination of these.

For understanding and tidying your data, it is paramount to identify the
types of cases in your data.

::: emphasis
What are you counting or plotting later on: persons, neighbourhoods,
persons-per-wave, ...?
:::

::: question
Inspect the `GlasgowFriends` tibble or use help on it to identify
different types of cases in this data set. For each variable, ask
yourself the question: What thing has this characteristic?
:::

```{r GlasgowFriends, exercise = TRUE}
str(GlasgowFriends)
?GlasgowFriends
```

### Primary key

Potential types of cases in `GlasgowFriends`:

-   **student**: has an age, sex, can be a friend, and so on.
-   **neighbourhood**: has a name, a distance from school.
-   **wave**: characteristic of neither a student nor a neighbourhood.

::: emphasis
Primary key = smallest set of variables for which every case (row) has a
unique combination of values.
:::

```{r}
#Example of observation missing for one student.
data.frame(
  student = c("s001", "s002", "s003", "s004"),
  age = c(13.3, 14.2, 13.5, 13.7),
  friendships = c(4, 2, 3, 6)
  ) |>
  knitr::kable(booktab = TRUE) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

The above data frame with students as cases:

-   has 1 row for each student;
-   variable `student` uniquely identifies a student;
-   `student` is the *primary key*.

### 

```{r}
#Example of observation missing for one student.
data.frame(
  student = c("s001", "s002", "s002", "s003"),
  age = c(13.3, 13.5, 14.2, 13.7),
  friendships = c(4, 2, 3, 6)
  ) |>
  knitr::kable(booktab = TRUE) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

::: question
What is the primary key in the above tibble? Are students the (only)
type of cases in this tibble?
:::

### The primary key can help to indicate the types of cases

Find the primary key of a data frame to understand the type(s) of cases.

Identifying the primary key of a data frame:

1.  From the data description, chose variables that you think identify
    types of cases.
2.  Count each unique combination of values on these variables with
    `count()`.
3.  Check if there are combinations of values that appear more then once
    with `filter(n > 1)`.
    -   If at least one combination of values appears more than once,
        this set of variables is not a primary key; add a variable to
        the key (go to Step 1).
    -   If no combination of values appears more than once, you may have
        more variables than the primary key. See if you can remove a
        variable and still have a unique key (go to Step 2).

::: question
Write code that identifies the primary key of tibble `GlasgowFriends`.
You may have to try out different combinations of variables; submit only
the final code for checking.
:::

```{r primarykeyhidden, eval=FALSE}
# primary key: this seems to be the only option with four variables
GlasgowFriends |> 
  count( student, wave, bestfriend, bfperiod ) |> 
  filter(n > 1)
```

```{r primarykey, exercise = TRUE}
# Add the variables in the count() function that you think 
# represent different types of cases (unit).
GlasgowFriends |> 
  count( ____ ) |> 
  filter(n > 1)
```

::: {#primarykey-hint}
**Hint:** Add the variables in the count() function that you think
represent different types of cases (unit).
:::

```{r primarykey-solution}
GlasgowFriends |> count( student, wave, bestfriend, bfperiod ) |> filter(n > 1)
```

```{r primarykey-check}
gradethis::grade_code(
  correct = "", 
  incorrect = " Well, maybe you just entered the variables in another order than I expected. If the result is 0 and you use just four variables, your have the correct primary key."
  )
```

### Finding the primary key

If you are unsure which variable to add to find the primary key:

-   Select cases with a non-unique combination of values on the
    variables you selected so far for the primary key.
-   Compare these cases on those variables that are NOT YET part of your
    work-in-progress primary key: Which variable has different values
    across cases?
-   This is the variable that you should add to get a primary key.

::: question
1.  Use the code below to find out which variable must be added because
    it has different values for the first pair of cases that have the
    same `student` and `wave` values.
2.  Add this variable to the code below. Any variable that must still be
    added to obtain a primary key?
:::

```{r primarykey2, exercise = TRUE}
# Display cases with the same provisional primary key as consecutive cases.
GlasgowFriends |>
  # (ADD HERE) for each combination of selected primary key variables...
  group_by( student, wave, bestfriend ) |>
  # ...calculate the number of cases with identical values on the grouping variables 
  mutate(n_identical = n()) |> 
  # retain only cases that are not unique on the selected variables (i.e. cases that identify this list of variables as incomplete primary key)
  filter(n_identical > 1) |>
  # (ALSO ADD HERE) sort the cases, so duplicates are next to each other
  arrange( student, wave, bestfriend ) |> 
  # show only the first two cases that are duplicates
  ungroup() |>
  slice(1:2)
```

### Identifying types of cases in a dataset

Every variable in your primary key and every combination of these
variables may represent a different type of case.

Example: A primary key consisting of the variables `student` and `wave`.

```{r}
data.frame(
  `Case type` = c("Student/person", "Wave", "Student per wave / wave per student"),
  Variables = c("student, neighbourhood, age, sex, ...", "wave ID", "alcohol, cannabis, tobacco, ..."),
  Notes = c("Fixed student characteristics", "Wave characteristics", "Student characteristics that may change between waves"),
  `Split tibble?` = c("yes", "no (only wave ID)", "yes")
  ) |>
  knitr::kable(booktab = TRUE,
               col.names = c("Case type", "Variables", "Notes", "Separate tibble?")
               ) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

### 

::: tip
**Tip**

Finding the variables that belong to a type of cases:

1.  Use common sense: The variable is a characteristic of what?
2.  Check that the variable has a unique value for each case of the
    selected type (use the code below), e.g., for each student or each
    student/wave combination.
3.  The variable belongs to the simplest type of cases (identified by
    fewest variables) satisfying the check in Step 2. For example,`age`
    is fixed per student and per student-wave combination.
:::

::: question
For each variable in `GlasgowFriends` determine to which type of case it
belongs.
:::

```{r casetype, exercise = TRUE}
# Code to check that a variable .var. always has the same value for each value
# of the selected type of case .type. . Replace .var. by a variable name and
# replace .type. by the variable(s) that define the type of case.
GlasgowFriends |>
  # For each value of the selected type of case...
  group_by( .type. ) |>
  # ... calculate the number of distinct values of the selected variable.
  summarise(n_distinct = n_distinct( .var. ), .groups = "drop") |>
  # Count number of distinct variable values per type of case: 
  # If 1, then variable has unique value for each case.
  count(n_distinct)
```

```{r casetype-hint-1}
# Example: Checking that variable age is unique for each student-wave combination.
GlasgowFriends |>
  # For each value of the selected type of case...
  group_by( student, wave ) |>
  # ... calculate the number of distinct values of the selected variable.
  summarise(n_distinct = n_distinct( age ), .groups = "drop") |>
  # Count number of distinct variable values per type of case (grouping has been
  # dropped): should always be 1.
  count(n_distinct)
# In the output of this code, 'n_distinct' gives the number of different ages
# encountered per student per wave. Column 'n' gives the number (frequency) of
# student per wave combinations.
```

```{r casetype-hint-2}
# If 'age' is unique for each student-wave combination, it can also be unique for each student. If so, 'age' belongs to the student type of case, not the student per wave type of case.
# Example: Checking that variable age is unique for each student-wave combination.
GlasgowFriends |>
  # For each value of the selected type of case...
  group_by( student ) |>
  # ... calculate the number of distinct values of the selected variable.
  summarise(n_distinct = n_distinct( age ), .groups = "drop") |>
  # Count number of distinct variable values per type of case (grouping has been dropped): should always be 1.
  count(n_distinct)
# In the output of this code, 'n_distinct' gives the number of different ages encountered per student. Column 'n' gives the number (frequency) of students.
```

### 

::: tip
**Programming Tip**

-   Be systematic and precise!
-   For example, if a primary key is defined as *the smallest set of
    variables for which every case has a unique combination of values*,
    check both parts of the definition:
    (1) Every case has a unique combination of values,
    (2) and this is the smallest set of variables for which 1 holds.
:::

### Splitting a tibble

::: emphasis
Tidy Rule 1. Each type of case must have its own tibble
:::

Create a separate tibble for each type of case.

This way, each piece of information is stored only once:

-   No notational variations, e.g., a person's ID number cannot be
    different in different cases.
-   Easy editing/updating: change/add a value only in one place.
-   Max storage efficiency: no repetition of information.
-   Information is stored by type of cases: easier to find.
-   Any multilevel structure of the data (for statistical analysis) is
    clear from the presence of different tibbles.

### 

::: question
Use your data wrangling skills from Session 2 and the data
transformation with dplyr cheat sheet to create a tibble for each type
of case in `GlasgowFriends`:

1.  Student.
:::

```{r studenttable, exercise = TRUE, exercise.lines = 6}
# Your code to create a tibble for student information without duplicates.
Student <- GlasgowFriends |> 
  # select the relevant variables
  select( ____ ) |> 
  # only keep cases that are distinct (not identical)
  distinct()
```

<!-- To hide the solution, use a textual hint. -->

::: {#studenttable-hint}
**Hint:** For now, include the neighbourhood variables in the Student
tibble.
:::

```{r studenttable-solution}
Student <- GlasgowFriends |> select(student, neighbourhood, schooldist, hoodname, age, sex, smoking_at_home, smoking_parents, smoking_siblings) |> distinct()
```

```{r studenttable-check}
gradethis::grade_result(
  pass_if(~ {nrow(.result) == 160 && ncol(.result) == 9 }, "Indeed, student, bestfriend, bfperiod, and bfwave belong together in this tibble."),
  fail_if(~ nrow(.result) != 160, "The number of cases in the new tibble is not correct. You did not select the correct variables."),
  fail_if(~ ncol(.result) != 9, "The number of variables in the new tibble is not correct. You did not select (all) correct variables.")
)
```

### 

::: question
2.  StudentWave.
:::

```{r studentwavetable, exercise = TRUE, exercise.lines = 6}
# Your code to create a tibble for student per wave information without duplicates.
StudentWave <- GlasgowFriends |> 
  # select the relevant variables
  select( ____ ) |> 
  # only keep cases that are distinct (not identical)
  distinct()
```

<!-- To hide the solution, use a textual hint. -->

::: {#studentwavetable-hint}
**Hint:** It is a good custom to use the primary key variables as the
first variables in a tibble.
:::

```{r studentwavetable-solution}
StudentWave <- GlasgowFriends |> select(student, wave, alcohol, cannabis, money, romantic, friend_1, friend_2, friend_3, friend_4, friend_5, friend_6 ) |> distinct()
```

```{r studentwavetable-check}
gradethis::grade_result(
  pass_if(~ {nrow(.result) == 480 && ncol(.result) == 12 }, "Indeed, student, bestfriend, bfperiod, and bfwave belong together in this tibble."),
  fail_if(~ nrow(.result) != 480, "The number of cases in the new tibble is not correct. You did not select the correct variables."),
  fail_if(~ ncol(.result) != 12, "The number of variables in the new tibble is not correct. You did not select (all) correct variables.")
)
```

### 

::: question
3.  StudentBestfriend.
:::

```{r bestfriendstable, exercise = TRUE, exercise.lines = 6}
# Your code to create a tibble for best friends information without duplicates.
StudentBestfriend <- GlasgowFriends |> 
  # select the relevant variables
  select( ____ ) |> 
  # only keep cases that are distinct (not identical)
  distinct()
```

<!-- To hide the solution, use a textual hint. -->

::: {#bestfriendstable-hint}
**Hint:** Include the `student` variable, otherwise we don't know who
nominated the best friend.
:::

```{r bestfriendstable-solution}
StudentBestfriend <- GlasgowFriends |> select(student, bestfriend, bfperiod, bfwave ) |> distinct()
```

```{r bestfriendstable-check}
gradethis::grade_result(
  pass_if(~ {nrow(.result) == 519 && ncol(.result) == 4 }, "Indeed, student, bestfriend, bfperiod, and bfwave belong together in this tibble."),
  fail_if(~ nrow(.result) != 519, "The number of cases in the new tibble is not correct. You did not select the correct variables."),
  fail_if(~ ncol(.result) != 4, "The number of variables in the new tibble is not correct. You did not select (all) correct variables.")
)
```

### 

Within the student tibble, the neighbourhood information is repeated:

-   Neighbourhood is a type of cases that should be split off into a
    separate tibble.
-   You know how to do that by now.

### Foreign key, one-to-many vs. many-to-many relations

::: emphasis
A foreign key is a variable in one table that is a primary key in
another tibble. It helps us to link cases in the two tables to each
other.
:::

-   Examples:
    -   `student` ("s001", "s002", ...) links the *Student* tibble to
        *StudentWave* and *StudentBestFriend*.
    -   `neighbourhood` links the *Student* and *Neighbourhood* tibbles.

```{r diagram0, out.width="60%", fig.asp=0.5, fig.cap="Figure. Relations between the tibbles. Variables that are (part of) the primary key of a tibble are underlined."}
library(DiagrammeR)
DiagrammeR::grViz("
digraph database {
    graph [rankdir = LR]
    node [
      shape=plaintext
      fontname = 'Helvetica'
    ]
    Student [label=<
<TABLE BORDER='0' CELLBORDER='1' CELLSPACING='0'>
  <TR><TD BGCOLOR='black'><FONT COLOR='white'>Student</FONT></TD></TR>
  <TR><TD PORT='f1'><U>student</U></TD></TR>
  <TR><TD PORT='f2'>neighbourhood</TD></TR>
  <TR><TD>age</TD></TR>
  <TR><TD>sex</TD></TR>
  <TR><TD>smoking_at_home</TD></TR>
  <TR><TD>smoking_parents</TD></TR>
  <TR><TD>smoking_siblings</TD></TR>
</TABLE>>];
    StudentBestfriend [label=<
<TABLE BORDER='0' CELLBORDER='1' CELLSPACING='0'>
  <TR><TD BGCOLOR='black'><FONT COLOR='white'>StudentBestfriend</FONT></TD></TR>
  <TR><TD PORT='f1'><U>student</U></TD></TR>
  <TR><TD><U>bestfriend</U></TD></TR>
  <TR><TD><U>bfperiod</U></TD></TR>
  <TR><TD>bfwave</TD></TR>
</TABLE>>];
    Neighbourhood [label=<
<TABLE BORDER='0' CELLBORDER='1' CELLSPACING='0'>
  <TR><TD BGCOLOR='black'><FONT COLOR='white'>Neighbourhood</FONT></TD></TR>
  <TR><TD PORT='f1'><U>neighbourhood</U></TD></TR>
  <TR><TD>schooldist</TD></TR>
  <TR><TD>hoodname</TD></TR>
</TABLE>>];
    StudentWave [label=<
<TABLE BORDER='0' CELLBORDER='1' CELLSPACING='0'>
  <TR><TD BGCOLOR='black'><FONT COLOR='white'>StudentWave</FONT></TD></TR>
  <TR><TD PORT='f1'><U>student</U></TD></TR>
  <TR><TD PORT='f2'><U>wave</U></TD></TR>
  <TR><TD>alcohol</TD></TR>
  <TR><TD>cannabis</TD></TR>
  <TR><TD>money</TD></TR>
  <TR><TD>romantic</TD></TR>
</TABLE>>];
    Neighbourhood:f1 -> Student:f2 ;
    Student:f1 -> StudentBestfriend:f1;
    Student:f1 -> StudentWave:f1
}
")
```

*Types of relations between tibbles:*

1.  *One-to-many relation*: Joining Student and Neighborhood tables on
    `neighborhood`. Each neighborhood occurs once in Neighborhood, but
    might occur multiple times in Students with multiple students living
    in the same neighborhoods.

2.  *Many-to-many relation*: Occurs when the foreign key - used to join
    two tables, is not a primary key in either table. Leads to untidy
    tables with lots of rows, but is sometimes necessary. (R will print
    a warning if you attempt a many-to-many join).

## 2. Each observation must have its own row.

For each student in each wave, up to six friends could be mentioned.

```{r showfriends}
GlasgowFriends |> 
  select(student, wave, friend_1, friend_2, friend_3, friend_4, friend_5, friend_6 ) |> 
  distinct() |>
  slice(1:6) |>
  kable(booktabs = TRUE) |>
  kableExtra::kable_classic(full_width = FALSE) |>
  kableExtra::column_spec(4, color = c("red", "black", "black", "red", "black", "black")) |>
  kableExtra::column_spec(5, color = c("black", "black", "red", "black", "black", "black"))
```

Problems:

-   Many cells with missing values (`NA`). Waste of space!
-   The same student may appear in each of the six friendship variables
    (e.g., `s028`). Hard to count!
-   The variable names contain a number (`friend_1`, `friend_2`, ...).
    But numbers are values (first mentioned friend, second mentioned
    friend, ...). How can we use these?

### 

### 

The tidy solution:

-   Stack repeated information.
-   Yields one row (observation) per student per wave per nominated
    friend.

```{r showfriendsgathered}
GlasgowFriends |> 
  select(student, wave, friend_1, friend_2, friend_3, friend_4, friend_5, friend_6 ) |>
  gather(key = "friendnumber", value = "friend", friend_1:friend_6, na.rm = TRUE) |>
  distinct() |>
  arrange(student, wave, friendnumber) |>
  slice(1:6) |>
  kable(booktabs = TRUE) |>
  kableExtra::kable_classic(full_width = FALSE) |>
  kableExtra::column_spec(4, color = c("black", "red", "black", "black", "black", "black"))
```

::: question
Use the tidyverse `pivot_longer()` function to create a new tibble named
`StudentWaveFriend` containing one row (observation) per student per
wave per nominated friend. The tibble should only contain the same
variables as in the above table.
:::

```{r gatherfriends, exercise = TRUE, exercise.setup = "studentwavetable-solution"}
StudentWaveFriend <- StudentWave |> 
  pivot_longer( ____ ) |>
  select()
#Show result.
StudentWaveFriend
```

::: {#gatherfriends-hint}
**Hint:** Use help on the pivot_longer() function .
:::

```{r gatherfriends-solution}
StudentWaveFriend <- StudentWave |> pivot_longer( friend_1:friend_6, names_to = "friendnumber", values_to = "friend", values_drop_na = TRUE ) |> select( student, wave, friendnumber, friend )
```

```{r gatherfriends-check}
gradethis::grade_result(
  fail_if(~ nrow(StudentWaveFriend) > 1600, "Did you remove the rows with missing values for the friend variable? These rows are superfluous. Use an argument in the `pivot_longer()` function to remove the rows."),
  fail_if(~ ("alcohol" %in% names(StudentWaveFriend)) | ("cannabis" %in% names(StudentWaveFriend)) | ("money" %in% names(StudentWaveFriend)) | ("romantic" %in% names(StudentWaveFriend)), "Drop the student characteristics that have only one value per student per wave: alcohol, cannabis, and so on. Why are they untidy in the new tibble?"),
  fail_if(~ ncol(StudentWaveFriend) < 4, "Did you select the relevant variables in the end: student, wave, and friend?"),
  pass_if(~ TRUE, "You retained the cases and variables that matter.")
)
```

### 

::: question
Oh, and remove the friendship data from tibble `StudentWave`.
:::

```{r gatherfriends2, exercise = TRUE, exercise.setup = "studentwavetable-solution"}
StudentWave <- StudentWave |> 
  ???
```

<div>

**Hint:** How about using the select() function?

</div>

```{r gatherfriends2-solution, eval=FALSE}
StudentWave <- StudentWave |> select(student:romantic)
```

```{r gatherfriends2-check}
gradethis::grade_result(
  fail_if(~ str_starts(names(StudentWave)[ncol(StudentWave)], "friend"), "There is still a friendship variable in the `StudentWave` tibble. Did you not overwrite this tibble?"),
  pass_if(~ ncol(StudentWave) == 6, "")
)
```

### Optional: `pivot_longer()` exercise

```{r pivotlonger-setup}
# create a tibble with waves as different columns
GlasgowWide <- UsingRTutorials::Glasgow |>
  pivot_wider(id_cols = student:smoking_siblings, names_from = wave, values_from = alcohol:friendships)
```

Tibble `GlasgowWide` contains the data in the original panel wave
format. (This tibble is loaded in this tutorial.)

::: question
Use the `pivot_longer()` function to gather the information measured per
wave in tibble `GlasgowWide`: smoking, alcohol, cannabis, and tobacco
use, pocket money, romantic relations, and number of friendships.
:::

Note: Send the gathered tibble to the screen, do not store it in a new
data object (so the code checker can do its work).

```{r pivotlonger, exercise = TRUE}
# Tip: Have a look at the data first.
GlasgowWide

```

::: {#pivotlonger-hint}
**Hint:** Read the vignette: Issue the R command `vignette("pivot")`.
Look for the part on multiple observations per row.
:::

```{r pivotlonger-solution}
GlasgowWide |> pivot_longer(cols = alcohol_t2:friendships_t1, names_to = c(".value", "wave"), names_sep = "_")
```

```{r pivotlonger-check}
gradethis::grade_code(
  correct = "", 
  incorrect = "Carefully read the part on multiple observations per row in the pivot vignette."
  )
```

## Fancy Stuff

### Networks

Many-to-many relations can be visualized as networks. Friendship
nominations among students can be depicted with nodes representing
students and arrows showing friendship nominations.

The `StudentWaveFriend` tibble contains all the information that we need
to draw a friendship network for the Glasgow data:

-   Variable `student` in this tibble represents the student nominating
    a friend.
-   Variable `friend` represents the student nominated.

A line in a network is just a pair of nodes!

There are many R packages for network visualization and network analysis
or modelling. [Network visualization with
R](https://kateto.net/network-visualization) offers a tutorial on some R
packages for network visualization. The examples below use the `igraph`
package.

::: question
Use the tutorial referenced above and the `igraph` documentation to play
around with the friendship network in wave1.
:::

```{r network1-setup}
#preparing data tibbles for network
Student <- GlasgowFriends |> select(student, neighbourhood, schooldist, hoodname, age, sex, smoking_at_home, smoking_parents, smoking_siblings) |> distinct()
StudentWave <- GlasgowFriends |> select(student, wave, alcohol, cannabis, money, romantic, friend_1, friend_2, friend_3, friend_4, friend_5, friend_6 ) |> distinct()
StudentWaveFriend <- StudentWave |> 
  gather(key = "friendnumber", value = "friend", friend_1:friend_6, na.rm = TRUE) |>
  select(student, wave, friend)
```

```{r network1, exercise = TRUE, exercise.lines = 20, exercise.eval = TRUE, fig.asp=0.5, out.width="100%", fig.cap="Figure: Friendship nominations at the first wave. Mutual nominations are black, unilateral nominations are orange."}
# Install the igraph package (once).
# install.packages("igraph")
# Load the igraph package.
library(igraph)
# Retrieve the friendship nominations in the first wave and ensure that the
# first two columns are the sender and receiver of the nomination.
arrows_wave1 <- StudentWaveFriend |>
  filter(wave == "t1") |>
  select(student, friend)
#add color variable for sexes
nodes_wave1 <- Student |>
  mutate(sex_color = ifelse(sex == "boy", "steelblue1", "tomato1"))
# Create an igraph object: a list of nodes (=vertices) combined with a list of lines/arrows.
net_wave1 <- igraph::graph_from_data_frame(d=arrows_wave1, vertices=nodes_wave1, directed=TRUE)

# Create a layout: positioning of the nodes.
set.seed(35546)
layout_1 <- igraph::layout_with_fr(net_wave1)
# Spread out the network horizontally.
layout_1 <- igraph::norm_coords(layout_1, ymin=-1, ymax=1, xmin=-2, xmax=2)
# Plot the network.
par(mar = c(0, 0, 0, 0))
plot(net_wave1, 
     vertex.label = NA, #hide node labels
     vertex.size = 8, #vertex size
     vertex.color = V(net_wave1)$sex_color,
     edge.arrow.size=.4, # arrowhead size
     #give reciprocal arrows (edges) a different colour than unilateral arrows:
     edge.color = ifelse(which_mutual(net_wave1), "gray40", "tan1"),
     rescale = FALSE, #don't rescale to square network
     layout = layout_1 #use the previously created node coordinates
     )
legend("topleft", c("Boys", "Girls"), pch=21, col="#777777", pt.bg=c("steelblue1", "tomato1"), pt.cex=2, cex=.8, bty="n", ncol=1)
```

### A geographic network

With `ggplot()` and the `map::` package, we can plot geographical
networks.

For example, the flights to/from NY City in the `nycflights13::`
(example data in the book).

```{r nycflightsmap, exercise = TRUE, exercise.eval = TRUE}
# Load the nycflights13 package.
library(nycflights13)
# Connect the flights (origin & destination) and airports tables (airport coordinates).
flights |> #use the flights tibble in the nycflights13 package
  #add coordinates to airport where flight originates
  inner_join(select(airports, origin = faa, origin_lat = lat, origin_lon = lon), 
             by = "origin") |>
  #add coordinates to destination airport
  inner_join(select(airports, dest = faa, dest_lat = lat, dest_lon = lon), by = "dest"
  ) |>
  # select the first 100 flights for illustrative purposes
  slice(1:100) |>
  # pipe the data into ggplot()
  ggplot(aes(
    x = origin_lon, xend = dest_lon,
    y = origin_lat, yend = dest_lat
    )) +
    borders("state") +
    geom_segment(arrow = arrow(length = unit(0.1, "cm"))) +
    coord_quickmap() +
    labs(y = "Latitude", x = "Longitude")
```

## What's next

-   We are half way through Sprint 2. Keep working on your Sprint
    Backlog.

-   Tips for your current or upcoming Sprint Backlog: With the skills
    discussed today, you can:

    -   Check and begin tidying your complex data sets according to
        principles 1 and 2.

    -   Summarize the contents and relations between your data sets in
        your Data Project Quarto file
