---
title: 'Session 9: Modelling 1'
output:
  learnr::tutorial:
    progressive: yes
    allow_skip: yes
runtime: shiny_prerendered
description: |
  Apply your skills in statistical analysis within R.
---

```{r setup, include=FALSE}
# Ensure that libraries are loaded.
library(tidyverse)
library(learnr)
library(gradethis)
library(knitr)
library(kableExtra)
library(haven) #For importing SPSS data files.
library(car) #For ANOVA.  
library(texreg) #For pretty regression results.
library(effects) #For two-way interaction plots.
library(broom) #For cleaning up statistical results.

tutorial_options(exercise.timelimit = 20, exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r, context="data", include=FALSE}
# Ensure that the data is loaded for the remainder of this tutorial.
flights4 <- UsingRTutorials::flights4
glbwarm <- UsingRTutorials::glbwarm
glbwarm_spss <- UsingRTutorials::glbwarm_spss
# The estimated regression model with rstanarm. 
model_1aBayes <- UsingRTutorials::model_1aBayes
```

<!-- Define programming tip style -->

```{=html}
<style>
.tip {
  background-color: #f5f5f5;
}
</style>
```
<!-- Define question style -->

```{=html}
<style>
.question {
  color: #5A9DDB;
}
</style>
```
<!-- Define emphasis style -->

```{=html}
<style>
.emphasis {
  color: #e8301b;
}
</style>
```
## Overview

-   t-test
-   Regression
-   Anova

## Basic Statistical Models

### 

Let us practice with some of the most common statistical analyses in R.

Consult Sections 3 and 4 in [*Help, My Collaborator Uses R! An Introduction to Reproducible Statistical Analyses in R*](https://wdenooy.github.io/Switch2R/index.html) and R help on the functions that we use.

### Example data

Example data: `glbwarm` (accessible within this tutorial).

Source: Erik Nisbet; <http://afhayes.com/>

Inspect the variables in the Environment.

Main data types: 1. Number: `govact`, `posemot`, `negemot`, `age`. 2. Character: `ideology`, `sex`, `partyid`.

::: question
Inspect variable summaries.
:::

```{r inspect, exercise = TRUE, exercise.eval = TRUE}
summary(glbwarm)
```

### *t* test: `t.test()`

Different types of *t* tests can be computed using one convenient formula.

Usage (in `?t.test`):

```{r t.test.vignette, eval=FALSE, echo=TRUE}
t.test(x, ...)

### Default S3 method:
t.test(x, y = NULL,
       alternative = c("two.sided", "less", "greater"),
       mu = 0, paired = FALSE, 
       var.equal = FALSE,
       conf.level = 0.95, ...)

### S3 method for class 'formula'
t.test(formula, data, subset, na.action, ...)
```

-   the function with just `x` is for a one sample *t* test: specify the hypothesized population mean with argument `mu =`.\
-   the function with `x, y` is for paired samples *t* tests.\
-   the function with a formula is for two samples *t* tests; `y` must be a variable with two categories.

#### `_` in a pipe

- All `tidyverse` functions have the data frame as the first argument.
- This is not necessarily the case for base R functions or other functions outside the `tidyverse`.
- If you use piping with these functions, use an underscore (`_`) as a placeholder to indicate where in the function the piped-in object should be entered.
- A second use of the `_` placeholder is when you do not use a function in the pipe.

**Example**

The code below tests whether the average arrival delay (`arr_delay0`) differs between the two airports of origin (`origin`) included in the `flights4` tibble and then extracts the t-statistic from the result. Pay attention to the placement of the `_`.

```{r pipedot, exercise = TRUE}
flights4 |> t.test(arr_delay0 ~ origin, data = _) |> #a function where data is not the first argument
  _$statistic #not a function
```

#### t-test Practice

::: question
Use a *t* test and the `glbwarm` data object for testing the following null hypotheses (in this order):

1.  Average negative emotions about global warming (variable `negemot`) are equal for females and males (variable `sex`) in the population.
2.  In the population, average negative emotions about global warming are 3.0.
3.  On average, negative emotions about global warming are higher than positive emotions about global warming (`posemot`).

Send the results to the screen.
:::

```{r ttest, exercise = TRUE}

```

```{r ttest-hint-1}
# Use the `t.test()` version that matches the kind of t test you need: on one
# mean, paired samples, or independent samples.
```

```{r ttest-hint-2}
# Note that the 'data = ' argument only works if we use the formula form 'y ~ x'.
# Independent samples t test:
glbwarm |> t.test(negemot ~ sex, data = _)
# For the other versions, the tibble name must be used and the dollar sign to
# fuly define the variable.
# t test on one mean (complete it yourself):
t.test(glbwarm$negemot, ... )
```

```{r ttest-hint-3}
# The code checker expects the three tests in the exact order as specified in
# the question.
```

```{r ttest-solution, exercise.reveal_solution = FALSE}
glbwarm |> t.test(negemot ~ sex, data = _)
t.test(glbwarm$negemot, mu = 3.0)
t.test(glbwarm$negemot, glbwarm$posemot, paired = TRUE)
```

```{r ttest-check}
gradethis::grade_code(
  correct = "", 
  incorrect = ""
  )
```

### Intermezzo: Extracting Statistical Results Using Lists

The summary of results given when you run a *t* test or another statistical test is great, but it is useful to be able to access individual elements of that result (e.g. the p- or T-value) separately, either to use them in calculations or present them in reproducible reports.

In R, the results of statistical tests are stored as lists. Let's briefly look at how to extract elements from a list.

#### Lists

### 

-   Vector: Collection of one type of elements.
-   List: Collection of (possibly) multiple types of elements.

A `List` may store anything (your perfect cupboard?)

Examples:

1.  Data frame (or tibble): list of (logic/num/char) vectors (variables).\
    Restriction: All vectors have the same length.\
2.  Statistical results: list with coefficients, etc.\
3.  Nested data: More than one value per variable per case.

#### Extracting statistical results

A *t* test yields a results object, which is a list.

```{r storeresults,exercise=TRUE, exercise.eval = TRUE}
# Store the results of a t test.
result_t <- glbwarm |> t.test(negemot ~ sex, data = _)
result_t #Default print method for these results.
```

-   Results of t test are saved as data object `result_t` here, so we can inspect the results object in the Environment (if you run this code in RStudio).

### 

Function `str()` shows the contents of a list.

::: question
Try to recognize the contents of this list.
:::

```{r str, exercise=TRUE, exercise.eval = TRUE, exercise.setup = "storeresults"}
str(result_t)
```

-   List elements are different: one number, two numbers, character string.\
-   "names" attribute: the name(s) of the list element value(s).
-   other attribute: other description of the list element value, e.g., the confidence level of the confidence interval.

Note:

-   *t* test results are stored as object of class: `htest` (bottom of the output);
-   Other R functions (e.g., `print()`) know what to do with the list.

#### Extracting an element from a list

We get elements from a list:

-   Appending the element name between `[[]]` to the list name;
-   Appending the element name with `$` to the list name.

::: question
Extract the confidence interval from the t test results data object (`result_t`) with `[[]]` and show (print) it (instead of saving it as a new data object).
:::

```{r extractbracket, exercise = TRUE, exercise.setup = "storeresults"}

```

<!-- To hide the solution, use a textual hint. -->

::: {#extractbracket-hint}
**Hint:** The element name must be in quotation marks. See the book section on *Recursive Vectors (Lists)*.
:::

```{r extractbracket-solution}
result_t[["conf.int"]]
```

```{r extractbracket-check}
gradethis::grade_code(
  correct = "", 
  incorrect = "Perhaps you used the element number in the list instead of the element name to extract the confidence level. That is OK."
  )
```

::: question
Do the same thing but now use `$`.
:::

```{r extractdollar, exercise = TRUE, exercise.setup = "storeresults"}

```

<!-- To hide the solution, use a textual hint. -->

::: {#extractdollar-hint}
**Hint:** See the book section on *Recursive Vectors (Lists)*.
:::

```{r extractdollar-solution}
result_t$conf.int
```

```{r extractdollar-check}
gradethis::grade_code(
  correct = "`$conf.int` is shorthand for `[['conf.int']]`.", 
  incorrect = ""
  )
```

### 

Let's go one level down in the list of t test results:

-   Extract the value of an attribute with the `attributes()` function.

```{r}
str(result_t[4])
```

::: question
Show the confidence level (as a number) using only dollar signs (`$`), not square brackets (`[[]]`). Show the result on the screen.
:::

```{r extractlevel, exercise = TRUE, exercise.setup = "storeresults"}

```

```{r extractlevel-hint-1}
# The confidence level is an attribute (`attr` in the structure overview presented above), so use the `attributes()` function.
attributes()
```

```{r extractlevel-hint-2}
# Pull out the confidence interval from the results. Complete this code.
attributes(result_t)
```

```{r extractlevel-hint-3}
# Have a look at the output of the attributes() function. How can you get the number of the confidence level?
attributes(result_t$conf.int)
```

```{r extractlevel-solution, exercise.reveal_solution = FALSE}
attributes(result_t$conf.int)$conf.level
```

```{r extractlevel-check}
gradethis::grade_code(
  correct = "We can use the dolar sign directly after the attributes() function. Isn't that nice?", 
  incorrect = "If you just get the number 0.95, you are fine. Probably, you used [[]] instead of $. If you get more than just the number, you have to go down one level in the list that `attributes()` generates."
  )
```

### 

Instead of the attribute value, we can get the attribute name with the `names()` function.

::: question
Extract the label "conf.level" from the list of t test results.
:::

```{r extractlabel, exercise = TRUE, exercise.setup = "storeresults"}

```

```{r extractlabel-hint-1}
# If you want the name of an attribute, apply the `names()` function to an attribute that you extract with attributes().
names(attributes( ))
```

```{r extractlabel-solution, exercise.reveal_solution = FALSE}
names(attributes(result_t$conf.int))
```

```{r extractlabel-check}
gradethis::grade_code(
  correct = "", 
  incorrect = "Did you use [[]] instead of $?"
  )
```

### 

Let us practice some more.

::: question
Get the name and value (rounded to one decimal place) of the mean of the second group in the results of the *t* test (`result_t`).
:::

```{r paste0, exercise = TRUE, exercise.setup = "storeresults"}

```

```{r paste0-hint-1}
# Build your code in steps. First, find the group means in the results.
str(result_t)
```

```{r paste0-hint-2}
# Second, pull the mean from the results data object. You know how to get only the second mean (complete the code below).
result_t$estimate
```

```{r paste0-hint-3}
# Third, round the mean with the round() function (use Help). Complete the code below.
round(result_t$estimate[[2]] )
```

```{r paste0-hint-4}
# Fourth, pull the name of the second group from the results data object. Complete the code below.
names(result_t$estimate)[2]
```

```{r paste0-solution, exercise.reveal_solution = FALSE}
names(result_t$estimate)[2]
round(result_t$estimate[[2]], digits = 1)
```

```{r paste0-check}
gradethis::grade_code(
  correct = "", 
  incorrect = "Check out the hints to this exercise. And mind the blanks in the resulting sentence."
  )
```

### 

::: tip
**Programming Tips**

-   Attribute names and attribute values are nested within a list element.\
-   `attributes()` creates a list; select from this list with \$ or [[]].
-   'Drill' into lower levels step by step:
    -   `attributes(result_t)` returns all first-level names and attributes (as a list);
    -   `attributes(result_t$conf.int)` returns all attributes of this element;
    -   `attributes(result_t$conf.int)$conf.level` gives the value of the conf.level attribute;\
    -   `names(attributes(result_t$conf.int))` returns the name of the conf.int attribute.\
-   `result_t$conf.int[1]` yields the lower limit of the confidence interval
:::

### F test on Two Variances: `var.test()`

Let's return to our *t* test. In contrast to SPSS, R only gives you what you ask for.

-   If you ask for a *t* test, you get a *t* test but not checks on assumptions.
-   You have to apply those checks yourself.

The version of the independent samples *t* test that we must use, depends on whether the population variances are equal for the two groups.

::: question
Use the function `var.test` to test if `govact` variances are equal for females and males in the population.

Use the `glbwarm` data object and store the results as a new data object named `vartest`.
:::

```{r vartest1, exercise = TRUE}
vartest <- ____
```

<!-- To hide the solution, use a textual hint. -->

::: {#vartest1-hint}
**Hint:** Have a look at the help for function `var.test`. It is important that you get used to the way R presents help on statistical functions.
:::

```{r vartest1-solution}
vartest <- var.test(govact ~ sex, data = glbwarm)
```

```{r vartest1-check}
gradethis::grade_code(
  correct = "", 
  incorrect = ""
  )
```

::: tip
**Remember**

-   R formula: dependent variable/outcome \~ independent variable/predictor (and more).
:::

### 

::: question
Pull the p value from data object `vartest` that you have just created.

Is the test on equal population variances statistically significant?
:::

```{r vartestp, exercise = TRUE, exercise.setup = "vartest1-solution"}

```

<!-- To hide the solution, use a textual hint. -->

::: {#vartestp-hint}
**Hint:** Review Session 5 if you do not know how to do this. Remember: function `str()` is handy to see the contents (structure) of a list.
:::

```{r vartestp-solution}
vartest$p.value
```

```{r vartestp-check}
gradethis::grade_code(
  correct = "`e-08`  (scientific notation) means `* 10^-8`, that is, divided by 10 to the power 8 (100,000,000). Note that the results are stored as class htest, just like the results from `t.test()`.", 
  incorrect = "Perhaps you used double square brackets instead of the dollar sign to pull out the p value. That's OK."
  )
```

### 

In R, we can use a function within an argument of another function.

Example for an independent samples *t* test:

-   `var-equal` argument is `FALSE` by default.
-   It must be `TRUE` if the *p* value of `var.test()` is larger then .05.

::: question
Integrate the F test on equal population variances in the *t* test, such that the *t* test automatically uses the correct version: with or without equal population variances assumed.

Send the results to the screen (do not save it as a data object).
:::

```{r ttest2, exercise = TRUE, exercise.setup = "vartest1-solution"}
t.test(govact ~ sex, data = glbwarm, var.equal = _____ )
```

```{r ttest2-hint-1}
# You already executed the t test in this tutorial. Add the var.equal argument.
t.test(govact ~ sex, data = glbwarm)
```

```{r ttest2-hint-2}
# In the preceding exercise, you pulled the p value from the stored test result.
vartest$p.value
# Add it to the var.equal argument in such a way that a p value over .05 yields TRUE.
```

```{r ttest2-hint-3}
# Replace the stored test result by the test function itself.
t.test(govact ~ sex, data = glbwarm, var.equal = vartest$p.value > 0.05)
```

```{r ttest2-solution, exercise.reveal_solution = FALSE}
t.test(govact ~ sex, data = glbwarm, var.equal = var.test(govact ~ sex, data = glbwarm)$p.value > 0.05)
```

```{r ttest2-check}
gradethis::grade_code(
  correct = "", 
  incorrect = ""
  )
```

### Linear Regression: `lm()`

Usage (in `?lm`):

```{r lm_vignette, eval=FALSE, echo=TRUE}
lm(formula, data, subset, weights, na.action,
   method = "qr", model = TRUE, x = FALSE, y = FALSE, qr = TRUE,
   singular.ok = TRUE, contrasts = NULL, offset, ...)
```

See (the [first edition](https://r4ds.had.co.nz/)) book p. 358-371 {Section 23.4} for using regression formulas: *In the second edition, modeling part has been removed.*

-   including interactions and
-   transformations within a formula.

::: question
Use `lm()` and tibble `glbwarm` to predict support for governmental action (`govact`) from age, negative emotions and party identification.

Store the results in data object `model_1`.
:::

```{r lm1, exercise = TRUE}

```

<!-- To hide the solution, use a textual hint. -->

::: {#lm1-hint}
**Hint:** `lm()` is not a tidyverse function, so you have to use the `data =` argument.

You can supply the name of the tibble (`glbwarm`) or pipe this tibble into the `lm()` function using the dot (`.`).
:::

```{r lm1-solution}
model_1 <- lm(govact ~ age + negemot + partyid, data = glbwarm)
```

```{r lm1-check}
gradethis::grade_code(
  correct = "", 
  incorrect = "Perhaps, you used the independent variables in a different order within the formula. That is fine."
  )
```

### 

For quick inspection, data objects for results of statistical analyses always have:

-   a `summary()` function;
-   a `print()` function .

Not for presentation of results!

::: question
Inspect the regression results (stored as `model_1`) with `summary()` and `print()`.

What happened to the character variable?
:::

```{r lm2, exercise = TRUE, exercise.setup = "lm1-solution"}

```

### Linear Regression: Two-Way Interaction

`lm()` takes care of:

-   creating dummies/indicator variables for a categorical predictor (character string or factor) - see preceding exercise;
-   creating interaction variables.

(This is easier than in SPSS.)

::: question
Add an interaction effect between negative emotions (numeric) and age (numeric, in decades) to the regression model.

Save the results as data object `model_1a`.

Show the results with `print()`. Can you interpret the interaction effect?
:::

```{r lm3, exercise = TRUE}

```

<!-- To hide the solution, use a textual hint. -->

::: {#lm3-hint}
**Hint:** An interaction term (`var1*var2`) in a regression formula yields the partial effects of the individual variables and their interaction effect(s).
:::

```{r lm3-solution}
model_1a <- lm(govact ~ age*negemot + partyid, data = glbwarm)
print(model_1a)
```

```{r lm3-check}
gradethis::grade_code(
  correct = "", 
  incorrect = "Perhaps, you used the independent variables in a different order within the formula. That is fine."
  )
```

### 

::: question
Now, predict support for governmental action (`govact`) from age and an interaction effect between negative emotions (numeric) and party identification (categorical).

Send the results to the screen. Can you make sense of the coefficients?
:::

```{r lm4, exercise = TRUE}

```

<!-- To hide the solution, use a textual hint. -->

::: {#lm4-hint}
**Hint:** R creates all dummy variables and all interaction variables. That is convenient!
:::

```{r lm4-solution}
lm(govact ~ age + negemot*partyid, data = glbwarm)
```

```{r lm4-check}
gradethis::grade_code(
  correct = "", 
  incorrect = ""
  )
```

### Analysis of Variance: `lm()` and `car::Anova()`

In R, analysis of variance consists of two steps.

**Step 1: ANOVA is linear regression with special contrasts (`contr.sum`).**

-   Contrast `contr.sum` gives deviations from the mean.
-   In analysis of variance, (main) effects are deviations from the (grand) mean.
-   The `contrasts =` argument requires:
    -   a list: `contrasts = list()`;
    -   with contrast type for each categorical predictor:
    -   `contrasts = list(sex = contr.sum, partyid = contr.sum)`

::: question
Estimate a regression model with support for governmental action (`govact`) predicted from respondent's sex and party identification, and the interaction between the two predictors.

Use `contr.sum` contrasts and save the results as data object `model_2`.
:::

```{r anova1, exercise = TRUE}
model_2 <- lm(govact ~ sex * partyid, data = glbwarm, contrasts= ____ )
```

<!-- To hide the solution, use a textual hint. -->

::: {#anova1-hint}
**Hint:** The `contrasts` argument requires a list of variable name and contrast type pairs.
:::

```{r anova1-solution}
model_2 <- lm(govact ~ sex * partyid, data = glbwarm, contrasts=list(sex=contr.sum, partyid=contr.sum))
```

```{r anova1-check}
gradethis::grade_code(
  correct = "Have a look at the results: send model_2 to the screen.", 
  incorrect = ""
  )
```

### 

**Step 2: Calculate the sums of squares partition.**

Functions:

-   `stats::anova()` for balanced designs.
-   `car::Anova()` for (balanced and) unbalanced designs (Type !! or III sums of squares).

::: question
Use the `Anova()` function to show the sums of squares partition with associated F tests of `model_2` on the screen.
:::

```{r anova2, exercise = TRUE, exercise.setup = "anova1-solution"}

```

<!-- To hide the solution, use a textual hint. -->

::: {#anova2-hint}
**Hint:** The `car` package has been loaded by the tutorial, so you do not have to include it if you use the `Anova()` function.
:::

```{r anova2-solution}
Anova(model_2)
```

```{r anova2-check}
gradethis::grade_code(
  correct = "", 
  incorrect = "Perhaps you used the package name in the command, which is fine."
  )
```

### 

The anova functions return a data frame, which you can use as any data frame.

For example, knit it to a pretty (HTML or PDF) table with `knitr::kable()`.

We will do that in the tutorial on Thursday.

### Missing Values

How a `stat::` package functions deal with missing values depends on the `na.action =` argument:

-   `na.omit` (default and preferred) or `na.exclude`: listwise deletion;
-   `na.fail`: stops with an error.

::: question
Check and, if necessary, set the `na.action` option in the console of RStudio.
:::

```{r NAoption, exercise = TRUE}
# Get the current option for na.action.
getOption("na.action")
# Set the option (if necessary).
options(na.action = "na.omit")
```

## Fancy Stuff

If you can execute regression models in R, you can also execute these using Bayesian statistics instead of traditional (frequentist) statistics.

The popularity of Bayesian statistics as an alternative to null hypothesis significance testing is growing. If you want to be among the first in your field going Bayesian, check out the short introduction provided in [Help, My Collaborator Goes Bayesian! Why And How To Apply Bayesian Data Analysis](https://wdenooy.github.io/Switch2Bayesian/index.html). Section 3.2 of the document offers a short introduction to using the `rstanarm` package for Bayesian data analysis.

The below code estimates a regression model predicting support for governmental action (`govact`) from age, negative emotions and party identification, with an interaction effect of age with negative emotions.

```{r rstanarm, echo=TRUE, eval=FALSE}
# Load the rstanarm package.
library(rstanarm)
# Estimate the regression model with rstanarm.
model_1aBayes <- rstanarm::stan_glm(govact ~ age * negemot + partyid,
                                    data = glbwarm)
# Standard output to screen.
print(model_1aBayes)
```

```{r rstanarm2, echo=FALSE, eval=TRUE}
# Shows only the output of print.
print(model_1aBayes)
```

Bayesian estimation yields a probability distribution for every parameter.

The printed summary gives the median of the probability distribution of a regression coefficient as its point estimate. In addition, it shows the Mean Average Deviation of the probability distribution: a simple type of standard deviation.

It is easy to get the posterior distributions as a data frame or tibble, so you can find any probability you like for a parameter.

```{r posterior1, exercise = TRUE, exercise.eval = TRUE}
# Extract the posteriors from the fitted model to a tibble.
posteriors <- as_tibble(model_1aBayes)
# Overview of variables in posteriors: each independent variable and the error term (sigma).
str(posteriors)
# Calculate the probability that the effect of negemot is larger than 0 in the population.
prob <- posteriors |>
  summarize(b_negemot = mean(negemot > 0))
# Plot the probability distribution and display this information.
posteriors |> 
  mutate(positive = negemot > 0) |>
  ggplot(mapping = aes(x = negemot)) +
    geom_histogram(
      aes(fill = positive),
      boundary = 0,
      bins = 30,
      show.legend = FALSE
      ) +
    geom_text(aes(label = prob$b_negemot, x = 0.15, y = 100))
```

The function `launch_shinystan()` in the `shinystan::` package (automatically loaded by package `rstanarm::`) offers an interactive overview of estimation and model checks and results.

```{r shinystan, echo = TRUE, eval=FALSE}
shinystan::launch_shinystan(model_1aBayes)
# Note: This function does not work from within a tutorial.
```

## What's next

-   We are in the middle of Sprint 5. Continue to work on your Sprint Backlog.

```{=html}
<!-- -->
```
-   Next session: Modelling 2/Presenting results
    -   Note: Statistical analyses are not necessary for the Data Project. The Data Project focuses on visualizations. You can, however, use statistical analysis to detect patterns in your data that you then try to visualize. Next session, we will discuss how to do this. Although R offers some off-the-shelf plots for presenting statistical results, it is usually better to create a custom plot that hopefully is more attractive and more informative than off-the-shelf statistical plots.
